---
output:
  pdf_document: default
  html_document: default
---
\newpage
\tableofcontents 
\newpage
\listoftables 
Table 1. Summary Statistics: Harmful Algal Bloom and Lake Characteristics of Lake Erie 2012 through 2018

Table 2. Summary Statistics: Harmful Algal Bloom and Lake Characteristics of Lake Erie at Bottom and Surface Depths

Table ###. Shapiro-Wilk Test for Normality Results

Table ###. Wilcoxin Rank Sum Test for Bottom versus Surface Depths

\newpage
\listoffigures 

Figure ###. Microcystin Concentrations at Surface and Bottom Depth for years 2015 through 2018

Figure ###. Chlorophyll a Concentrations at Surface and Bottom Depth for years 2015 through 2018

Figure ###. Dissolved Oxygen Concentrations at Surface and Bottom Depth for years 2015 through 2018

\newpage

```{r setup, include=FALSE}
# Set your working directory
getwd()

# Load your packages
library(tidyverse)
#install.packages("lubridate")
library(lubridate)
library(ggplot2)
#install.packages("gt")
library(gt)
#install.packages("pixiedust")
library(pixiedust)

# Set your ggplot theme
ggplot_theme <- theme_linedraw(base_size = 12) +
  theme(panel.grid=element_line(colour="gray"),
        axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(ggplot_theme)

# Load your datasets
Erie_Data_2012_2018 <- 
  read.csv("./Data/Raw/lake_erie_habs_field_sampling_results_2012_2018_v2.csv",
           check.names = F)

#Erie_Data_2012_2018 <-   read.csv("./Data/Processed/Erie_2012_2018_processed.csv",
      #    check.names = F)
```

```{r, include = FALSE}
#Clean up column names
colnames(Erie_Data_2012_2018) <- gsub(" ", ".", colnames(Erie_Data_2012_2018))
colnames(Erie_Data_2012_2018) <- gsub("/", ".", colnames(Erie_Data_2012_2018))
colnames(Erie_Data_2012_2018) <- gsub("[//+]", "", colnames(Erie_Data_2012_2018))
colnames(Erie_Data_2012_2018) <- gsub("[()]", "", colnames(Erie_Data_2012_2018))
colnames(Erie_Data_2012_2018) <- gsub("[<>]", "", colnames(Erie_Data_2012_2018))

colnames(Erie_Data_2012_2018)

colnames(Erie_Data_2012_2018)[19] <- "Photo.active.radiation"
```

```{r, include = FALSE}
#Create new data frame 
Erie_Data_2012_2018 <- data_frame(Erie_Data_2012_2018$Date,
                                  Erie_Data_2012_2018$Site,
                                  Erie_Data_2012_2018$Sample.Depth.category,
                                  Erie_Data_2012_2018$Dissolved.Organic.Carbon.mg.L,
                                  Erie_Data_2012_2018$Urea.b5g.N.L,
                                  Erie_Data_2012_2018$Nitrate..Nitrite.mg.N.L,
                                  Erie_Data_2012_2018$Ammonia.b5g.N.L,
                                 Erie_Data_2012_2018$Soluble.Reactive.Phosphorus.b5g.P.L,
                                  Erie_Data_2012_2018$Total.Dissolved.Phosphorus.b5g.P.L,
                                 Erie_Data_2012_2018$Total.Phosphorus.b5g.P.L,
                                  Erie_Data_2012_2018$Extracted.Chlorophyll.a.b5g.L,
                                  Erie_Data_2012_2018$Extracted.Phycocyanin.b5g.L,
                                  Erie_Data_2012_2018$Dissolved.Microcystin.b5g.L,
                                  Erie_Data_2012_2018$CTD.Dissolved.Oxygen.mg.L,
                                  Erie_Data_2012_2018$Photo.active.radiation,
                                  Erie_Data_2012_2018$Sample.Temperature.b0C,
                                  Erie_Data_2012_2018$CTD.Temperature.b0C)

colnames(Erie_Data_2012_2018) <- gsub("Erie_Data_2012_2018", "", colnames(Erie_Data_2012_2018))

colnames(Erie_Data_2012_2018) <- gsub("[//$]", "", colnames(Erie_Data_2012_2018))
```

```{r, include = FALSE}
#change character variables to numeric; change date column to date format
lapply(Erie_Data_2012_2018, class)
Erie_Data_2012_2018$Dissolved.Microcystin.b5g.L <- as.numeric(Erie_Data_2012_2018$Dissolved.Microcystin.b5g.L)
Erie_Data_2012_2018$Extracted.Phycocyanin.b5g.L <- as.numeric(Erie_Data_2012_2018$Extracted.Phycocyanin.b5g.L)
Erie_Data_2012_2018$Soluble.Reactive.Phosphorus.b5g.P.L <- as.numeric(Erie_Data_2012_2018$Soluble.Reactive.Phosphorus.b5g.P.L)
Erie_Data_2012_2018$Ammonia.b5g.N.L <- as.numeric(Erie_Data_2012_2018$Ammonia.b5g.N.L)
Erie_Data_2012_2018$Nitrate..Nitrite.mg.N.L <- as.numeric(Erie_Data_2012_2018$Nitrate..Nitrite.mg.N.L)
Erie_Data_2012_2018$Urea.b5g.N.L <- as.numeric(Erie_Data_2012_2018$Urea.b5g.N.L)
Erie_Data_2012_2018$Date <- as.Date(Erie_Data_2012_2018$Date, format = "%m/%d/%Y")

#remove scum
Erie_Data_2012_2018 <- Erie_Data_2012_2018 %>%
  filter(!Sample.Depth.category=="Scum")

#Add a year column
Erie_Data_2012_2018 <- Erie_Data_2012_2018 %>% 
  mutate(Year = year(Date))
```

```{r, include=FALSE}
#write.csv(Erie_Data_2012_2018, "./Data/Processed/Erie_2012_2018_2_processed.csv")
```


```{r, include=FALSE}
#Filter for surface and bottom variables
bottom_surface_Erie <- Erie_Data_2012_2018 %>% 
  filter(Sample.Depth.category == "Bottom" | Sample.Depth.category == "Surface")

#filter for years 2015 through 2018; no bottom observations are collected before 2015
bottom_surface_Erie <- bottom_surface_Erie %>% 
  filter(Year == "2015" | Year == "2016" | Year == "2017" | Year == "2018")

```

```{r, include=FALSE}
##2022 data
Erie_Data_2022 <- read.csv("./Data/Raw/2022_WLE_Weekly_Datashare_CSV.csv")

##changing character columns to numeric
class(Erie_data$Temp_C)
Erie_data$Temp_C <- as.numeric(Erie_data$Temp_C)

class(Erie_data$DO_mgL.1)
Erie_data$DO_mgL.1 <- as.numeric(Erie_data$DO_mgL.1)

class(Erie_data$Dissolved_Microcystin_ugL.1)
Erie_data$Dissolved_Microcystin_ugL.1 <- 
  as.numeric(Erie_data$Dissolved_Microcystin_ugL.1)

class(Erie_data$Extracted_CHLa_ugL.1)

class(Erie_data$Extracted_PC_ugL.1)
Erie_data$Extracted_PC_ugL.1 <- as.numeric(Erie_data$Extracted_PC_ugL.1)

##creating subset
Erie_data_subset <- data_frame(Erie_data$Date, Erie_data$Sample_Depth_category, 
                               Erie_data$Temp_C, Erie_data$DO_mgL.1,
                          Erie_data$Dissolved_Microcystin_ugL.1,
                          Erie_data$Extracted_CHLa_ugL.1, 
                          Erie_data$Extracted_PC_ugL.1)

##changing column names
colnames(Erie_data_subset) <- c("Date", "Depth_category", "Temp", "DO", "MC", "Chla", "PC")
View(Erie_data_subset)

##filtering out scum
Erie_data_subset <- Erie_data_subset %>%
  filter(!Depth_category=="Scum")

##creating processed dataset
write.csv(Erie_data_subset, "./Data/Processed/Erie_2022_processed.csv")
Erie_data_processed <- read.csv("./Data/Processed/Erie_2022_processed.csv")
#View(Erie_data_processed)

```


# Rationale and Research Questions

Harmful algal blooms (HABs) are characterized by rapid algae population growth coupled with toxin production in an aquatic system (NOAA, 2016). In recent history HABs have been shown to be increasing in prevalence with documented observations throughout the U.S. and the world (U.S. National Office for HABs, n.d.). This increased prevalence of HABs poses a threat to both public and ecosystem health (U.S. EPA, 2022a). Generally, HABs are associated with an influx of environmental nutrients into an aquatic system; however, other regional or lake-specific variables such as temperature, physical chemical properties, and lake structure can influence the development of HABs (CDC, 2022). Although HABs have been largely stated to have increased through time, many of these variables are challenging to generalize across water bodies due to the complexity of algal bloom formation. It is for this reason that this report will focus on HAB development in one water body: Lake Erie.  

Lake Erie has been affected by seasonal HABs since the 1990s (NSF, 2019). Lake Erie is adjacent to multiple metropolitan areas with populations that exceed 50,000 people and provides drinking water for 12 million people (U.S. EPA, 2022b). In 2015 Lake Erie had a bloom which covered over 300 square miles, thus making it one of the largest algal blooms documented in recent history (LEF, n.d.). Algae and toxin production has public health implications for surrounding populations as exposure can occur through recreation and consumption of contaminated drinking water (Dierkes, 2014; U.S. EPA, 2022b). Algal blooms can also create anoxic conditions when algae undergo decomposition in the environment which also influences ecosystem health (CDC, 2022). Due to the historical and current public health and ecosystem prevalence, data collection efforts have been implemented within Lake Erie to aid with forecasting future bloom severity. These data in the western basin are evaluated in this report.

Lake Erie is the shallowest of the great lakes with an average depth of 19 meters (m); the western portion, which comprises approximately 20% of the lake, has an average depth of 7.4 m and a maximum depth of 19 m (U.S. EPA, 2022b). Western Lake Erie is expected to undergo stratification for a short period of time during the summer months, leading to a warmer surface layer (epilimnion) and cooler bottom layer (hypolimnion; U.S. EPA, 2022b). Nutrient inputs come from a variety of sources which include wastewater treatment plants and agriculture which can influence algal bloom production (Dean, 2022). Microcystis is the most common species and microcystin, a potent hepatotoxin, is the most common toxin documented in HABs in Lake Erie. Considering the characteristics of Western Lake Erie, this report evaluates the following questions:

  What are the drivers of harmful algal blooms in western Lake Erie?
  
  1. Which variables (i.e., temperature, nitrate/nitrite, and/or phosphorus) contribute to microcystin production, Microcystis population growth, and dissolved oxygen?
  
  2. How does Microcystis and microcystin concentration change over time?
  
  3. Are there differences between toxin and algae concentrations at different lake depths?  



\newpage

# Dataset Information

The NOAA Great Lakes Environmental Research Laboratory (NOAA GLERL) is a federal research laboratory designed to monitor and research the ecology and hydrology of the Great Lakes Region. The NOAA GLERL provides data critical to studying the presence and harmful impacts of algal blooms in the great lakes region.

The NOAA GLERL studies the movement, size, and concentration of toxins and nutrients typically associated with HABs, such as microcystin, nitrogen, phycocyanin, and chlorophyll. They also measure other factors that impact HABs, such as temperature and Dissolved oxygen (DO) concentrations. The lab’s primary goal is to characterize the growth and abundance of microcystin concentrations and how their presence affects the toxicity of the lake.

The data from Western Lake Erie comes from nine sites sampled weekly, four of which have buoys collecting data continuously at 15-minute intervals. Only some sampled areas gather data on the same parameters, resulting in occasional missing data. 

The sourced data set contains information on parameters ranging from 2012 - 2018 and a final data set that monitors the toxin and nutrient concentration from 2022.






\newpage

# Exploratory Analysis 

```{r}
##histograms here
```


To begin exploration of the questions, a summary statistics table (Table 1) was generated to show the number of observations, mean, standard deviation, and median within each year for all variables of interest (i.e., microcystin concentration, chlorophyll a concentration, dissolved oxygen concentration, nitrate/nitrite concentration, dissolved phosphorus, and temperature. This was conducted to evaluate balance within the data regarding the number of observations but also to get a glimpse into trends occuring within given years. 

***Table 1. Summary Statistics: Harmful Algal Bloom and Lake Characteristics of Lake Erie 2012 through 2018***

```{r, include = FALSE, message = FALSE, warning = FALSE}
#Load processed dataset
Erie_Data_2012_2018 <- read.csv("./Data/Processed/Erie_2012_2018_2_processed.csv")

# Select Variables of Interest
Erie.select<-Erie_Data_2012_2018%>%
 dplyr::select(Year, Dissolved.Microcystin.b5g.L, Extracted.Chlorophyll.a.b5g.L, CTD.Dissolved.Oxygen.mg.L, Nitrate..Nitrite.mg.N.L, Total.Dissolved.Phosphorus.b5g.P.L, CTD.Temperature.b0C)
head(Erie.select)

```

```{r, include = FALSE, message = FALSE, warning = FALSE}

Erie.select.longer<-pivot_longer(Erie.select, !Year, names_to = "Variable", values_to = "value")

summary.Erie<-Erie.select.longer%>%
  dplyr::group_by(Year, Variable) %>%
  dplyr::summarize(
    Obs=length(value)-sum(is.na(value)),
    Mean=mean(value, na.rm=TRUE),
    SD=sd(value, na.rm=TRUE),
    Median=median(value, na.rm=TRUE))

#Adding column for units; maybe make separate table with just variables of interest and units?
summary.Erie$Unit <- c("milligrams/Liter", "Celsius", "micrograms/Liter", "micrograms/Liter", "milligrams of Nitrogen/Liter", "micrograms of Phosphorus/Liter", "milligrams/Liter", "Celsius", "micrograms/Liter", "micrograms/Liter", "milligrams of Nitrogen/Liter", "micrograms of Phosphorus/Liter", "milligrams/Liter", "Celsius", "micrograms/Liter", "micrograms/Liter", "milligrams of Nitrogen/Liter", "micrograms of Phosphorus/Liter", "milligrams/Liter", "Celsius", "micrograms/Liter", "micrograms/Liter", "milligrams of Nitrogen/Liter", "micrograms of Phosphorus/Liter", "milligrams/Liter", "Celsius", "micrograms/Liter", "micrograms/Liter", "milligrams of Nitrogen/Liter", "micrograms of Phosphorus/Liter", "milligrams/Liter", "Celsius", "micrograms/Liter", "micrograms/Liter", "milligrams of Nitrogen/Liter", "micrograms of Phosphorus/Liter", "milligrams/Liter", "Celsius", "micrograms/Liter", "micrograms/Liter", "milligrams of Nitrogen/Liter", "micrograms of Phosphorus/Liter")
   
head(summary.Erie)

#remove columns which have NAs
summary.Erie_dropNAs <- summary.Erie %>% 
  drop_na()
  
```

```{r, include = FALSE, message = FALSE, warning = FALSE}
#recode
summary.Erie_dropNAs$Variable <- dplyr:: recode(summary.Erie_dropNAs$Variable, 
                                         Dissolved.Microcystin.b5g.L = "Dissolved Microcystin (ug/L)",
                                         Extracted.Chlorophyll.a.b5g.L="Chlorophyll a (ug/L)",
                                         CTD.Dissolved.Oxygen.mg.L="Dissolved Oxygen (mg/L)",
                                         Nitrate..Nitrite.mg.N.L="Nitrate/Nitrite Concentration (mg/L)",
                                         Total.Dissolved.Phosphorus.b5g.P.L="Dissolved Phosphorus (ug/L",
                                         CTD.Temperature.b0C= "Temperature (deg C)")
```

```{r, warning = FALSE, meassage = FALSE, echo = FALSE}
#Create Table

summary.Erie_dropNAs %>%
gt() %>%
 


  cols_width(
    c(Variable)~ px(120),
    c(Obs)~ px(120),
    everything() ~ px(90))%>%

  
  
  tab_header(
    title = md("Summary Statistics: Lake Erie Harmful Algal Bloom Characteristics in 2012 through 2018")) %>%
  
 
  fmt_passthrough (columns=c(Variable)) %>%
  fmt_number(columns = c(Obs), decimals = 0) %>%
  fmt_number(columns = c(Median), decimals = 2) %>%
  fmt_number(columns = c(Mean), decimals = 2) %>%
  fmt_number(columns = c(SD), decimals = 2) %>%
  

cols_label(
    Variable = "",
    Obs = "Oberservations",
    Median = "Median",
    Mean = "Mean",
    SD = "Standard Deviation")
  

```

To explore the third question, a summary statistics table (Table 2) was generated to show the number of observations, mean, standard deviation, and median within each group of interest (i.e., surface and bottom depth categories). This was conducted to evaluate balance within the data regarding the number of observations but also to get a glimpse into trends for both bottom and surface depth observations. 

***Table 2. Summary Statistics: Harmful Algal Bloom and Lake Characteristics of Lake Erie at Bottom and Surface Depths***
```{r, include = FALSE, message = FALSE, warning = FALSE}
# Select Variables of Interest

Erie.select<-bottom_surface_Erie%>%
 dplyr::select(Sample.Depth.category, Dissolved.Microcystin.b5g.L, Extracted.Chlorophyll.a.b5g.L, CTD.Dissolved.Oxygen.mg.L, Nitrate..Nitrite.mg.N.L, Total.Dissolved.Phosphorus.b5g.P.L, CTD.Temperature.b0C)
head(Erie.select)

```

```{r, include = FALSE, message = FALSE, warning = FALSE}

Erie.select.longer<-pivot_longer(Erie.select, !Sample.Depth.category, names_to = "Variable", values_to = "value")

summary.Erie<-Erie.select.longer%>%
  dplyr::group_by(Sample.Depth.category, Variable) %>%
  dplyr::summarize(
    Obs=length(value)-sum(is.na(value)),
    Mean=mean(value, na.rm=TRUE),
    SD=sd(value, na.rm=TRUE),
    Median=median(value, na.rm=TRUE))
   
head(summary.Erie)

#remove columns which have NAs
summary.Erie_dropNAs <- summary.Erie %>% 
  drop_na()
  
```

```{r, include = FALSE, message = FALSE, warning = FALSE}
#recode
summary.Erie_dropNAs$Variable <- dplyr:: recode(summary.Erie_dropNAs$Variable, 
                                         Dissolved.Microcystin.b5g.L = "Dissolved Microcystin (ug/L)",
                                         Extracted.Chlorophyll.a.b5g.L="Chlorophyll a (ug/L)",
                                         CTD.Dissolved.Oxygen.mg.L="Dissolved Oxygen (mg/L)",
                                         Nitrate..Nitrite.mg.N.L="Nitrate/Nitrite Concentration (mg/L)",
                                         Total.Dissolved.Phosphorus.b5g.P.L="Dissolved Phosphorus (ug/L",
                                         CTD.Temperature.b0C= "Temperature (deg C)")
```

```{r, warning = FALSE, meassage = FALSE, echo = FALSE}
#Create Table

summary.Erie_dropNAs %>%
gt() %>%
 


  cols_width(
    c(Variable)~ px(120),
    c(Obs)~ px(120),
    everything() ~ px(90))%>%

  
  
  tab_header(
    title = md("Summary Statistics: Lake Erie Harmful Algal Bloom Characteristics at Bottom and Surface Depths")) %>%
  
 
  fmt_passthrough (columns=c(Variable)) %>%
  fmt_number(columns = c(Obs), decimals = 0) %>%
  fmt_number(columns = c(Median), decimals = 2) %>%
  fmt_number(columns = c(Mean), decimals = 2) %>%
  fmt_number(columns = c(SD), decimals = 2) %>%
  

cols_label(
    Variable = "",
    Obs = "Oberservations",
    Median = "Median",
    Mean = "Mean",
    SD = "Standard Deviation")
  

```


\newpage

# Analysis



## Question 1: Which variables (i.e., temperature, nitrate/nitrite, and/or phosphorus) contribute to microcystin production, Microcystis population growth, and dissolved oxygen?

## Question 2: How does Microcystis and microcystin concentration change over time?

## Question 3: Are there differences between toxin and algae concentrations at different lake depths?

Regarding the third question it is hypothesized that both microcystin, chlorophyll a, and dissolved oxygen concentrations will be greater at surface depths opposed to bottom depths in Lake Erie. These hypotheses are stated below:

***Hypotheses for microcystin concentration at surface vs. bottom depth***

Ho: Surface depth microcystin concentrations are <= bottom depth microcystin concentrations

HA: Surface depth microcystin concentrations are > bottom depth microcystin concentrations

***Hypotheses for chlorophyll a concentration at surface vs. bottom depth***

Ho: Surface depth Chlorophyll a concentrations are <= bottom depth microcystin concentrations

HA: Surface depth Chlorophyll a concentrations are > bottom depth microcystin concentrations

***Hypotheses for dissolved oxygen concentration at surface vs. bottom depth***

Ho: Surface depth dissolved oxygen concentrations are <= bottom depth microcystin concentrations

HA: Surface depth dissolved oxygen concentrations are > bottom depth microcystin concentrations

Before assessing these claims, a Shapiro-Wilk test for normality was conducted for the microcystin, chlorophyll a, and dissolved oxygen data. The test rejected the null hypothesis which states that the data are normally distributed in all cases as shown in Table ###. In result, the Wilcoxon Rank Sum test was employed to evaluate the means between surface and bottom layers for each variable. These results can be found in Table ###. 

```{r, include=FALSE}
#test for normality of data
shapiro.microcystin <- shapiro.test(bottom_surface_Erie$Dissolved.Microcystin.b5g.L) #the data is not shown to be normal; reject the null hypothesis of normality in this case

shapiro.Chla <- shapiro.test(bottom_surface_Erie$Extracted.Chlorophyll.a.b5g.L) #the data is not shown to be normal; reject the null hypothesis of normality in this case

shapiro.DO <- shapiro.test(bottom_surface_Erie$CTD.Dissolved.Oxygen.mg.L)#the data is not shown to be normal; reject the null hypothesis of normality in this case
```

```{r, include=FALSE}
#Shapiro-wilk test table formatting
shapiro.1 <- rbind(shapiro.microcystin, shapiro.Chla, shapiro.DO)
shapiro<- as.data.frame(shapiro.1)

shapiro <- shapiro %>% 
  select(!method)

colnames_data <- c("Dissolved Microcystin", "Chlorophyll a", "Dissolved Oxygen")

shapiro$data.name <- 
  replace(shapiro$data.name,values = colnames_data)

shapiro.10 <- shapiro %>% 
  mutate_at(vars(1,2),  funs(gsub("[c()W=]", "", .)))

shapiro.10_num <- shapiro.10 %>% 
  mutate_if(is.character, as.numeric)

shapiro.10_round <- shapiro.10_num %>% 
  mutate_at(vars(1,2),  funs(round(.,digits = 3)))

colnames(shapiro.10_round) <- c('Statistic','P-Value','Data')

shapiro_reorder <- shapiro.10_round[, c(3, 1, 2)]

```

***Table ###. Shapiro-Wilk Test for Normality Results***

```{r, warning = FALSE, meassage = FALSE, echo = FALSE}
gt(shapiro_reorder) %>% 
  tab_header(
    title = "Shapiro-Wilk Test Results")
```

```{r, include = FALSE}
#Drop NAs

bottom_surface_Erie_dropNA_microcystin <- bottom_surface_Erie %>% 
  drop_na(Dissolved.Microcystin.b5g.L)

bottom_surface_Erie_dropNA_CHLa <- bottom_surface_Erie %>% 
  drop_na(Extracted.Chlorophyll.a.b5g.L)

bottom_surface_Erie_dropNA_DO <- bottom_surface_Erie %>% 
  drop_na(CTD.Dissolved.Oxygen.mg.L)
```

```{r, include = FALSE}
#Microcystin: surface vs bottom

wilcox_microcystin <- wilcox.test(Dissolved.Microcystin.b5g.L ~ Sample.Depth.category, data = bottom_surface_Erie_dropNA_microcystin, alternative = "less") #Due to nonnormality of the data, a rank sum test was applied

wilcox_microcystin

#CHLa: surface vs bottom

wilcox_chla <- wilcox.test(Extracted.Chlorophyll.a.b5g.L ~ Sample.Depth.category, data = bottom_surface_Erie_dropNA_CHLa, alternative = "less") #Due to nonnormality of the data, a rank sum test was applied

wilcox_chla

wilcox_DO <- wilcox.test(CTD.Dissolved.Oxygen.mg.L ~ Sample.Depth.category, data = bottom_surface_Erie_dropNA_DO, alternative = "less") #Due to nonnormality of the data, a rank sum test was applied

wilcox_DO
```

```{r, include=FALSE}
#Wilcoxin Rank Sum test table formatting
wilcox.1 <- rbind(wilcox_microcystin, wilcox_chla, wilcox_DO)
wilcox<- as.data.frame(wilcox.1)

wilcox <- wilcox %>% 
  select(statistic, p.value, data.name)

colnames_data_w <- c("Dissolved Microcystin", "Chlorophyll a", "Dissolved Oxygen")

wilcox$data.name <- 
  replace(wilcox$data.name,values = colnames_data_w)

wilcox.10 <- wilcox %>% 
  mutate_at(vars(1,2),  funs(gsub("[c()W=]", "", .)))

wilcox.10_num <- wilcox.10 %>% 
  mutate_if(is.character, as.numeric)

wilcox.10_round <- wilcox.10_num %>% 
  mutate_at(vars(1,2),  funs(round(.,digits = 3)))

colnames(wilcox.10_round) <- c('Statistic','P-Value','Data')

wilcox_reorder <- wilcox.10_round[, c(3, 1, 2)]

```
***Table ###. Wilcoxin Rank Sum Test for Bottom versus Surface Depths***

```{r, warning = FALSE, meassage = FALSE, echo = FALSE}
gt(wilcox_reorder) %>% 
  tab_header(
    title = "Wilcoxin Rank Sum Test Results")
```

When looking at microcystin concentrations at surface and bottom depths, the Wilcoxin Rank Sum test showed a nonsignificant result and thus a failure to reject the null hypothesis which states that microcystin concentrations at surface depth are less than or equal to microcystin concentrations at bottom depth (W = 24800, p-value = 0.1753). Figure ### shows these data. When looking at chlorophyll a concentrations at surface and bottom depths, the Wilcoxin Rank Sum test showed a significant result and a rejection of the null hypothesis which states that chlorophyll a concentrations at surface depth are less than or equal to chlorophyll a concentrations at bottom depth (W = 78212, p-value = 0.0008634). Figure ### shows these data. Lastly, when looking at dissolved oxygen concentrations, the Wilcoxin Rank Sum test showed a significant result and a rejection of the null hypothesis which states that dissolved oxygen concentrations at surface depth are less than or equal to dissolved concentrations at bottom depth (W = 51924, p-value < 2.2e-16). Figure ### shows these data.

```{r, warning = FALSE, meassage = FALSE, echo = FALSE}
#Microcystin boxplot
Microcystin_SB_plot <- ggplot(bottom_surface_Erie_dropNA_microcystin)+ 
  geom_boxplot(aes(x = Year, y=Dissolved.Microcystin.b5g.L, fill = Sample.Depth.category))+
  ylim(0,0.75)+
  labs(y = "Microcystin Concentration (ug/L)", x = "Sample Depth Category", title = "Microcystin Concentration at Surface and Bottom Depths")

print(Microcystin_SB_plot)
```
***Figure ###. Microcystin Concentrations at Surface and Bottom Depth for years 2015 through 2018***

```{r, warning = FALSE, meassage = FALSE, echo = FALSE}
#CHLa boxplot: surface vs bottom
CHLa_SB_plot <- ggplot(bottom_surface_Erie_dropNA_CHLa)+ 
  geom_boxplot(aes(x = Year, y=Extracted.Chlorophyll.a.b5g.L, fill = Sample.Depth.category))+
  ylim(0,100)+
  labs(y = "Extracted Chlorophyll a Concentration (ug/L)", x = "Sample Depth Category", title = "Extracted Chlorophyll a Concentration at Surface and Bottom Depths")

print(CHLa_SB_plot)
```

***Figure ###. Chlorophyll a Concentrations at Surface and Bottom Depth for years 2015 through 2018***

```{r, warning = FALSE, meassage = FALSE, echo = FALSE}
#DO boxplot: surface vs bottom
DO_SB_plot <- ggplot(bottom_surface_Erie_dropNA_DO)+ 
  geom_boxplot(aes(x = Year, y=CTD.Dissolved.Oxygen.mg.L, fill = Sample.Depth.category))+
  labs(y = "Dissolved Oxygen Concentration (mg/L)", x = "Sample Depth Category", title = "Dissolved Oxygen Concentration at Surface and Bottom Depths")

print(DO_SB_plot)
```

***Figure ###. Dissolved Oxygen Concentrations at Surface and Bottom Depth for years 2015 through 2018***



\newpage

# Summary and Conclusions


\newpage

# References
<add references here if relevant, otherwise delete this section> 

Centers for Disease Control and Prevention (CDC; 2022). Causes and Ecosystem Impacts. https://www.cdc.gov/habs/environment.html#:~:text=Levels%20of%20nutrients%20such%20as%20phosphorus%20and%20nitrogen%20increase%20in,Sewage%20from%20people%20and%20animals

Dean, S. (2022). Harmful algal blooms in Lake Erie expected to be smaller this year, says NOAA. https://www.michigan.gov/egle/newsroom/mi-environment/2022/08/11/harmful-algal-blooms-in-lake-erie-expected-to-be-smaller-this-year-says-noaa#:~:text=The%20Lake%20Erie%20HABs%20consist,treatment%20of%20their%20source%20water.
Dierkes, C. (2014). Algae Through the Ages. https://ohioseagrant.osu.edu/news/2014/17gmc/algae-through-the-ages

Lake Erie Foundation (LEF; n.d.) Harmful Algal Blooms. https://lakeeriefoundation.org/issues/nutrientsharmful-algae/#:~:text=Harmful%20algal%20blooms%20(HABs)%20are,and%20the%20resulting%20algal%20blooms.

National Oceanic and Atmospheric Administration (NOAA; 2016). What is a harmful algal bloom? https://www.noaa.gov/what-is-harmful-algal-bloom

National Science Foundation (NSF; 2019). Lake Erie's toxic algae blooms: Why is the water turning green? https://beta.nsf.gov/news/lake-eries-toxic-algae-blooms-why-water-turning#:~:text=Since%20the%20late%201990s%2C%20Lake,warnings%20to%20avoid%20water%20contact. 

United States Environmental Protection Agency (U.S. EPA; 2022a). Harmful Algal Blooms. https://www.epa.gov/nutrientpollution/harmful-algal-blooms

United States Environmental Protection Agency (U.S. EPA; 2022b). Lake Erie. https://www.epa.gov/greatlakes/lake-erie
U.S. National Office for HABs. Recent Trends: National Changes. (n.d.) https://hab.whoi.edu/maps/regions-us-distribution/regions-us-recent-trends/
```

