---
output:
  pdf_document: default
  html_document: default
---
\newpage
\tableofcontents 

\newpage
\listoftables 
Table 1. Summary Statistics: Harmful Algal Bloom and Lake Characteristics of Lake Erie 2012 through 2018

Table 2. Summary Statistics: Harmful Algal Bloom and Lake Characteristics of Lake Erie at Bottom and Surface Depths

Table ###. Shapiro-Wilk Test for Normality Results

Table ###. Wilcoxin Rank Sum Test for Bottom versus Surface Depths

\newpage
\listoffigures 
Figure 1. Chlorophyll a distribution from 2012-2018.

Figure 2. Chlorophyll a distribution in 2022.

Figure 3. Temperature distribution from 2012-2018 (top) and in 2022 (bottom).

Figure 4. Microcystin distribution from 2012-2018 (top) and in 2022 (bottom).

Figure 5. Dissolved oxygen distribution from 2012-2018 (top) and in 2022 (bottom).

Figure ###. Microcystin Concentrations at Surface and Bottom Depth for years 2015 through 2018

Figure ###. Chlorophyll a Concentrations at Surface and Bottom Depth for years 2015 through 2018

Figure ###. Dissolved Oxygen Concentrations at Surface and Bottom Depth for years 2015 through 2018

\newpage

```{r setup, include=FALSE}
# Set your working directory
getwd()

# Load your packages
library(tidyverse)
#install.packages("lubridate")
library(lubridate)
library(ggplot2)
#install.packages("gt")
library(gt)
#install.packages("pixiedust")
library(pixiedust)
#install.packages("broom")
library(broom)

# Set your ggplot theme
ggplot_theme <- theme_linedraw(base_size = 12) +
  theme(panel.grid=element_line(colour="gray"),
        axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(ggplot_theme)

# Load your datasets
Erie_Data_2012_2018 <- 
  read.csv("./Data/Raw/lake_erie_habs_field_sampling_results_2012_2018_v2.csv",
           check.names = F)

```

```{r, include=FALSE}
#Clean up column names
colnames(Erie_Data_2012_2018) <- gsub(" ", ".", colnames(Erie_Data_2012_2018))
colnames(Erie_Data_2012_2018) <- gsub("/", ".", colnames(Erie_Data_2012_2018))
colnames(Erie_Data_2012_2018) <- gsub("[//+]", "", colnames(Erie_Data_2012_2018))
colnames(Erie_Data_2012_2018) <- gsub("[()]", "", colnames(Erie_Data_2012_2018))
colnames(Erie_Data_2012_2018) <- gsub("[<>]", "", colnames(Erie_Data_2012_2018))

colnames(Erie_Data_2012_2018)

colnames(Erie_Data_2012_2018)[19] <- "Photo.active.radiation"
```

```{r, include=FALSE}
#Create new data frame 
Erie_Data_2012_2018 <- data_frame(Erie_Data_2012_2018$Date,
                                  Erie_Data_2012_2018$Site,
                               Erie_Data_2012_2018$Sample.Depth.category,                          Erie_Data_2012_2018$Dissolved.Organic.Carbon.mg.L,
Erie_Data_2012_2018$Urea.b5g.N.L,
Erie_Data_2012_2018$Nitrate..Nitrite.mg.N.L,
Erie_Data_2012_2018$Ammonia.b5g.N.L,
Erie_Data_2012_2018$Soluble.Reactive.Phosphorus.b5g.P.L,
Erie_Data_2012_2018$Total.Dissolved.Phosphorus.b5g.P.L,
Erie_Data_2012_2018$Total.Phosphorus.b5g.P.L,
Erie_Data_2012_2018$Extracted.Chlorophyll.a.b5g.L,
Erie_Data_2012_2018$Extracted.Phycocyanin.b5g.L,
Erie_Data_2012_2018$Dissolved.Microcystin.b5g.L,
Erie_Data_2012_2018$CTD.Dissolved.Oxygen.mg.L,
Erie_Data_2012_2018$Photo.active.radiation,
Erie_Data_2012_2018$Sample.Temperature.b0C,
Erie_Data_2012_2018$CTD.Temperature.b0C)

colnames(Erie_Data_2012_2018) <- gsub("Erie_Data_2012_2018", "", colnames(Erie_Data_2012_2018))

colnames(Erie_Data_2012_2018) <- gsub("[//$]", "", colnames(Erie_Data_2012_2018))
```

```{r, include=FALSE}
#change character variables to numeric; change date column to date format
lapply(Erie_Data_2012_2018, class)
Erie_Data_2012_2018$Dissolved.Microcystin.b5g.L <- as.numeric(Erie_Data_2012_2018$Dissolved.Microcystin.b5g.L)
Erie_Data_2012_2018$Extracted.Phycocyanin.b5g.L <- as.numeric(Erie_Data_2012_2018$Extracted.Phycocyanin.b5g.L)
Erie_Data_2012_2018$Soluble.Reactive.Phosphorus.b5g.P.L <- as.numeric(Erie_Data_2012_2018$Soluble.Reactive.Phosphorus.b5g.P.L)
Erie_Data_2012_2018$Ammonia.b5g.N.L <- as.numeric(Erie_Data_2012_2018$Ammonia.b5g.N.L)
Erie_Data_2012_2018$Nitrate..Nitrite.mg.N.L <- as.numeric(Erie_Data_2012_2018$Nitrate..Nitrite.mg.N.L)
Erie_Data_2012_2018$Urea.b5g.N.L <- as.numeric(Erie_Data_2012_2018$Urea.b5g.N.L)
Erie_Data_2012_2018$Date <- as.Date(Erie_Data_2012_2018$Date, format = "%m/%d/%Y")

#remove scum
Erie_Data_2012_2018 <- Erie_Data_2012_2018 %>%
  filter(!Sample.Depth.category=="Scum")

```

```{r, include=FALSE}
#write.csv(Erie_Data_2012_2018, "./Data/Processed/Erie_2012_2018_2_processed.csv")
```


```{r, include=FALSE}
#Filter for surface and bottom variables
bottom_surface_Erie.1 <- Erie_Data_2012_2018 %>% 
  filter(Sample.Depth.category == "Bottom" | Sample.Depth.category == "Surface")

#Add year column
bottom_surface_Erie <- bottom_surface_Erie.1 %>% 
  mutate(Year = format(Date, "%Y"))

#filter for years 2015 through 2018; no bottom observations are collected before 2015
bottom_surface_Erie <- bottom_surface_Erie %>% 
  filter(Year == "2015" | Year == "2016" | Year == "2017" | Year == "2018")

```

```{r, include=FALSE}
##2022 data
Erie_Data_2022 <- read.csv("./Data/Raw/2022_WLE_Weekly_Datashare_CSV.csv")

##changing character columns to numeric
class(Erie_Data_2022$Temp_C)
Erie_Data_2022$Temp_C <- as.numeric(Erie_Data_2022$Temp_C)

class(Erie_Data_2022$DO_mgL.1)
Erie_Data_2022$DO_mgL.1 <- as.numeric(Erie_Data_2022$DO_mgL.1)

class(Erie_Data_2022$Dissolved_Microcystin_ugL.1)
Erie_Data_2022$Dissolved_Microcystin_ugL.1 <- 
  as.numeric(Erie_Data_2022$Dissolved_Microcystin_ugL.1)

class(Erie_Data_2022$Extracted_CHLa_ugL.1)

class(Erie_Data_2022$Extracted_PC_ugL.1)
Erie_Data_2022$Extracted_PC_ugL.1 <- as.numeric(Erie_Data_2022$Extracted_PC_ugL.1)

##creating subset
Erie_data_subset <- data_frame(Erie_Data_2022$Date, Erie_Data_2022$Sample_Depth_category, 
                               Erie_Data_2022$Temp_C, Erie_Data_2022$DO_mgL.1,
                          Erie_Data_2022$Dissolved_Microcystin_ugL.1,
                          Erie_Data_2022$Extracted_CHLa_ugL.1, 
                          Erie_Data_2022$Extracted_PC_ugL.1)

##changing column names
colnames(Erie_data_subset) <- c("Date", "Depth_category", "Temp", "DO", "MC", "Chla", "PC")
View(Erie_data_subset)

##filtering out scum
Erie_data_subset <- Erie_data_subset %>%
  filter(!Depth_category=="Scum")

##creating processed dataset
#write.csv(Erie_data_subset, "./Data/Processed/Erie_2022_processed.csv")
Erie_Data_2022 <- read.csv("./Data/Processed/Erie_2022_processed.csv")
#View(Erie_Data_2022)

```


# Rationale and Research Questions

Harmful algal blooms (HABs) are characterized by rapid algae population growth coupled with toxin production in an aquatic system (NOAA, 2016). In recent history HABs have been shown to be increasing in prevalence with documented observations throughout the U.S. and the world (U.S. National Office for HABs, n.d.). This increased prevalence of HABs poses a threat to both public and ecosystem health (U.S. EPA, 2022a). Generally, HABs are associated with an influx of environmental nutrients into an aquatic system; however, other regional or lake-specific variables such as temperature, physical chemical properties, and lake structure can influence the development of HABs (CDC, 2022). Although HABs have been largely stated to have increased through time, many of these variables are challenging to generalize across water bodies due to the complexity of algal bloom formation. It is for this reason that this report will focus on HAB development in one water body: Lake Erie.  

Lake Erie has been affected by seasonal HABs since the 1990s (NSF, 2019). Lake Erie is adjacent to multiple metropolitan areas with populations that exceed 50,000 people and provides drinking water for 12 million people (U.S. EPA, 2022b). In 2015 Lake Erie had a bloom which covered over 300 square miles, thus making it one of the largest algal blooms documented in recent history (LEF, n.d.). Algae and toxin production has public health implications for surrounding populations as exposure can occur through recreation and consumption of contaminated drinking water (Dierkes, 2014; U.S. EPA, 2022b). Algal blooms can also create anoxic conditions when algae undergo decomposition in the environment which also influences ecosystem health (CDC, 2022). Due to the historical and current public health and ecosystem prevalence, data collection efforts have been implemented within Lake Erie to aid with forecasting future bloom severity. These data in the western basin are evaluated in this report.

Lake Erie is the shallowest of the great lakes with an average depth of 19 meters (m); the western portion, which comprises approximately 20% of the lake, has an average depth of 7.4 m and a maximum depth of 19 m (U.S. EPA, 2022b). Western Lake Erie is expected to undergo stratification for a short period of time during the summer months, leading to a warmer surface layer (epilimnion) and cooler bottom layer (hypolimnion; U.S. EPA, 2022b). Nutrient inputs come from a variety of sources which include wastewater treatment plants and agriculture which can influence algal bloom production (Dean, 2022). Microcystis is the most common species and microcystin, a potent hepatotoxin, is the most common toxin documented in HABs in Lake Erie. Considering the characteristics of Western Lake Erie, this report evaluates the following questions:

  What are the drivers of harmful algal blooms in western Lake Erie?
  
  1. How does Microcystis and microcystin concentration change over time?
  
  2. Which variables (i.e., temperature, nitrate/nitrite, and/or phosphorus) contribute to microcystin production, Microcystis population growth, and dissolved oxygen?
  
  3. Are there differences between toxin and algae concentrations at different lake depths?  



\newpage

# Dataset Information

The National Oceanic and Atmospheric Administration Great Lakes Environmental Research Laboratory (NOAA GLERL) is a federal research laboratory designed to monitor and research the ecology and hydrology of the Great Lakes Region. The NOAA GLERL provides data critical to studying the presence and harmful impacts of algal blooms in the great lakes region.

The NOAA GLERL studies the movement, size, and concentration of toxins and nutrients typically associated with HABs, such as microcystin, nitrogen, phycocyanin, and chlorophyll. They also measure other factors that impact HABs, such as temperature and Dissolved oxygen (DO) concentrations. The lab’s primary goal is to characterize the growth and abundance of microcystin concentrations and how their presence affects the toxicity of the lake.

The data from Western Lake Erie comes from nine sites sampled weekly, four of which have buoys collecting data continuously at 15-minute intervals. Only some sampled areas gather data on the same parameters, resulting in occasional missing data. 

The sourced data set contains information on parameters ranging from 2012 - 2018 and a final data set that monitors the toxin and nutrient concentration from 2022. The variables we will analyze are displayed below with their units of measurement. 

***Table ?. Variables of interest and associated units***
```{r, warning = FALSE, message = FALSE, echo = FALSE}
#Create table: variables of interest and units
#Create dataframe 
data_df_2012_2018 <- data_frame(
  Variable = c("Dissolved Oxygen", "Temperature", "Dissolved Microcystin", "Chlorophyll a", "Nitrate + Nitrite", "Total Dissolved Phosphorus"),
  Unit = c("milligrams/Liter", "Celsius", "micrograms/Liter", "micrograms/Liter", "milligrams of Nitrogen/Liter", "micrograms of Phosphorus/Liter"))
#View(data_df_2012_2018)

#Create table
data_df_2012_2018 %>%
gt() %>%
  cols_width(
    c(Variable)~ px(120),
    c(Unit)~ px(150),
    everything() ~ px(90))%>%
  
  tab_header(
    title = md("Variables of Interest")) %>%

  fmt_passthrough (columns=c(Variable)) %>%
  fmt_passthrough(columns = c(Unit)) %>%

cols_label(
    Variable = "",
    Unit = "Unit") %>%

cols_align(
    align = "center")

```


\newpage

# Exploratory Analysis 

We created histograms to illustrate the distribution of our variables of interest. Chlorophyll a values from 2012-2018 are largely concentrated below 100 $\mu g/L$ (Fig. 1). We discovered an outlier from August 2015, when chlorophyll a was measured at 6784 $\mu$g/L. [sentence about how it was a historic algal bloom year] We removed this outlier in the second histogram to better display the distribution of data.

```{r, warning = FALSE, message = FALSE, echo = FALSE}
#Load processed 2012-2018 data
Erie_Data_2012_2018 <- read.csv("./Data/Processed/Erie_2012_2018_processed.csv")

#Histograms showing distribution of variables 
#Chlorophyll a: 2012-2018
chla_hist_2012_2018 <- ggplot(data=Erie_Data_2012_2018, aes(x=Extracted.Chlorophyll.a.b5g.L))+
  geom_histogram(bins=50, color="black", fill="lightgray") +
  xlab(expression("Chlorophyll a" ~ "("*mu*"g/L)"))+
  ggtitle("2012-2018")
chla_hist_2012_2018

chla_hist_no_outlier <- ggplot(data=subset(Erie_Data_2012_2018, !Extracted.Chlorophyll.a.b5g.L>6000), aes(x=Extracted.Chlorophyll.a.b5g.L))+
  geom_histogram(bins=50, color="black", fill="lightgray") +
  xlab(expression("Chlorophyll a" ~ "("*mu*"g/L)"))+
  ggtitle("2012-2018")
chla_hist_no_outlier
##easier to see distribution of chl a when outlier is removed

```
***Figure 1. Chlorophyll a distribution from 2012-2018. Bottom figure excludes 6784 $\mu$g/L$ outlier from August, 2015.***

Similarly, chlorophyll a measurements from 2022 are clustered below 100 $\mu g/L$, with few observations between 150 and 272 $\mu g/L$ (Fig. 2). 

```{r, warning = FALSE, message = FALSE, echo = FALSE}
#Chlorophyll a: 2022
chla_hist_2022 <- ggplot(data=Erie_Data_2022, aes(x=Chla))+
  geom_histogram(bins=50, color="black", fill="lightgray") +
  xlab(expression("Chlorophyll a" ~ "("*mu*"g/L)"))+
  ggtitle("2022")
chla_hist_2022

```
***Figure 2. Chlorophyll a distribution in 2022.***

Temperature measurements in both the 2012-2018 and 2022 datasets predominantly fall between 20 and 25 degrees Celsius (Fig. 3).

```{r, warning = FALSE, message = FALSE, echo = FALSE}
#Temperature: 2012-2018
temp_hist_2012_2018 <- ggplot(data=Erie_Data_2012_2018, aes(x=CTD.Temperature.b0C))+
  geom_histogram(bins=50, color="black", fill="lightgray") +
  xlab(expression("Temperature" ~ "("*degree*C*")"))+
  ggtitle("2012-2018")
temp_hist_2012_2018

#Temperature: 2022
temp_hist_2022 <- ggplot(data=Erie_Data_2022, aes(x=Temp))+
  geom_histogram(bins=50, color="black", fill="lightgray")+
  xlab(expression("Temperature" ~ "("*degree*C*")"))+
  ggtitle("2022")
temp_hist_2022

```
***Figure 3. Temperature distribution from 2012-2018 (top) and in 2022 (bottom).***

Dissolved microcystin concentrations between 2012 and 2018 were strongly right-skewed, with the majority of observations falling below 0.25 $\mu g/L$ and very few between 1.3 and 2 $\mu g/L$. Microcystin measurements in 2022 are mainly below 0.5 $\mu g/L$ and exhibit a narrower range of values (Fig. 4).

```{r, warning = FALSE, message = FALSE, echo = FALSE}
#Microcystin: 2012-2018
mc_hist_2012_2018 <- ggplot(data=Erie_Data_2012_2018, aes(x=Dissolved.Microcystin.b5g.L))+
  geom_histogram(bins=50, color="black", fill="lightgray")+
  xlab(expression("Microcystin" ~ "("*mu*"g/L)"))+
  ggtitle("2012-2018")
mc_hist_2012_2018

#Microcystin: 2022
mc_hist_2022 <- ggplot(data=Erie_Data_2022, aes(x=MC))+
  geom_histogram(bins=50, color="black", fill="lightgray")+
  xlab(expression("Microcystin" ~ "("*mu*"g/L)"))+
  ggtitle("2022")
mc_hist_2022

```
***Figure 4. Microcystin distribution from 2012-2018 (top) and in 2022 (bottom).***

Dissolved oxygen measurements from 2012-2018 are concentrated around 8 mg/L, while 2022 seems to have overall lower dissolved oxygen values, mainly concentrated between 7 and 8 mg/L (Figure 5).

```{r, warning = FALSE, message = FALSE, echo = FALSE}
#Dissolved oxygen: 2012-2018
do_hist_2012_2018 <- ggplot(data=Erie_Data_2012_2018, aes(x=CTD.Dissolved.Oxygen.mg.L))+
  geom_histogram(bins=50, color="black", fill="lightgray")+
  xlab("Dissolved Oxygen (mg/L)") +
  ggtitle("2012-2018")
do_hist_2012_2018

#Dissolved oxygen: 2022
do_hist_2022 <- ggplot(data=Erie_Data_2022, aes(x=DO))+
  geom_histogram(bins=50, color="black", fill="lightgray")+
  xlab("Dissolved Oxygen (mg/L)") +
  ggtitle("2022")
do_hist_2022

```
***Figure 5. Dissolved oxygen distribution from 2012-2018 (top) and in 2022 (bottom).***


To begin exploration of the questions, a summary statistics table (Table 1) was generated to show the number of observations, mean, standard deviation, median, maximum, and minimum values within each year for all variables of interest (i.e., microcystin concentration, chlorophyll a concentration, dissolved oxygen concentration, nitrate/nitrite concentration, dissolved phosphorus, and temperature. This was conducted to evaluate balance within the data regarding the number of observations but also to get a glimpse into trends occuring within given years. 

***Table 1. Summary Statistics: Harmful Algal Bloom and Lake Characteristics of Lake Erie 2012 through 2018***

```{r, include = FALSE, message = FALSE, warning = FALSE}
#Convert Date column to date format
Erie_Data_2012_2018$Date <- as.Date(Erie_Data_2012_2018$Date, format="%m/%d/%Y")

#Add year column
Erie_Data_2012_2018 <- Erie_Data_2012_2018 %>%
  mutate(Year = year(Date))

# Select variables of interest
Erie.select<-Erie_Data_2012_2018%>%
 dplyr::select(Year, Dissolved.Microcystin.b5g.L, Extracted.Chlorophyll.a.b5g.L, CTD.Dissolved.Oxygen.mg.L, Nitrate..Nitrite.mg.N.L, Total.Dissolved.Phosphorus.b5g.P.L, CTD.Temperature.b0C)
head(Erie.select)

```

```{r, include = FALSE, message = FALSE, warning = FALSE}

Erie.select.longer<-pivot_longer(Erie.select, !Year, names_to = "Variable", values_to = "value")

summary.Erie<-Erie.select.longer%>%
  dplyr::group_by(Year, Variable) %>%
  dplyr::summarize(
    Obs=length(value)-sum(is.na(value)),
    Mean=mean(value, na.rm=TRUE),
    SD=sd(value, na.rm=TRUE),
    Median=median(value, na.rm=TRUE),
    Max = max(value, na.rm=TRUE),
    Min = min(value, na.rm = TRUE))
   
head(summary.Erie)

#remove columns which have NAs
summary.Erie_dropNAs <- summary.Erie %>% 
  drop_na()
  
```

```{r, include = FALSE, message = FALSE, warning = FALSE}
#recode
summary.Erie_dropNAs$Variable <- dplyr:: recode(summary.Erie_dropNAs$Variable, 
                                         Dissolved.Microcystin.b5g.L = "Dissolved Microcystin (ug/L)",
                                         Extracted.Chlorophyll.a.b5g.L="Chlorophyll a (ug/L)",
                                         CTD.Dissolved.Oxygen.mg.L="Dissolved Oxygen (mg/L)",
                                         Nitrate..Nitrite.mg.N.L="Nitrate/Nitrite Concentration (mg/L)",
                                         Total.Dissolved.Phosphorus.b5g.P.L="Dissolved Phosphorus (ug/L)",
                                         CTD.Temperature.b0C= "Temperature (deg C)")

```

```{r, warning = FALSE, message = FALSE, echo = FALSE}
#Create Table
summary.Erie_dropNAs %>%
gt() %>%
 
  cols_width(
    c(Variable)~ px(120),
    c(Obs)~ px(120),
    everything() ~ px(90))%>%

  tab_header(
    title = md("Summary Statistics: Lake Erie Harmful Algal Bloom Characteristics in 2012 through 2018")) %>%

  fmt_passthrough (columns=c(Variable)) %>%
  fmt_number(columns = c(Obs), decimals = 0) %>%
  fmt_number(columns = c(Median), decimals = 2) %>%
  fmt_number(columns = c(Mean), decimals = 2) %>%
  fmt_number(columns = c(SD), decimals = 2) %>%
  fmt_number(columns = c(Max), decimals = 2) %>% 
  fmt_number(columns = c(Min), decimals = 2) %>% 

cols_label(
    Variable = "",
    Obs = "Oberservations",
    Median = "Median",
    Mean = "Mean",
    SD = "St Deviation",
    Max = "Max",
    Min="Min")

```

\newpage
To explore the third question, a summary statistics table (Table 2) was generated to show the number of observations, mean, standard deviation, and median within each group of interest (i.e., surface and bottom depth categories). This was conducted to evaluate balance within the data regarding the number of observations but also to get a glimpse into trends for both bottom and surface depth observations. 

***Table 2. Summary Statistics: Harmful Algal Bloom and Lake Characteristics of Lake Erie at Bottom and Surface Depths***
```{r, include = FALSE, message = FALSE, warning = FALSE}
# Select Variables of Interest

Erie.select<-bottom_surface_Erie%>%
 dplyr::select(Sample.Depth.category, Dissolved.Microcystin.b5g.L, Extracted.Chlorophyll.a.b5g.L, CTD.Dissolved.Oxygen.mg.L, Nitrate..Nitrite.mg.N.L, Total.Dissolved.Phosphorus.b5g.P.L, CTD.Temperature.b0C)
head(Erie.select)

```

```{r, include = FALSE, message = FALSE, warning = FALSE}

Erie.select.longer<-pivot_longer(Erie.select, !Sample.Depth.category, names_to = "Variable", values_to = "value")

summary.Erie<-Erie.select.longer%>%
  dplyr::group_by(Sample.Depth.category, Variable) %>%
  dplyr::summarize(
    Obs=length(value)-sum(is.na(value)),
    Mean=mean(value, na.rm=TRUE),
    SD=sd(value, na.rm=TRUE),
    Median=median(value, na.rm=TRUE),
    Max = max(value, na.rm=TRUE),
    Min = min(value, na.rm = TRUE))
   
head(summary.Erie)

#remove columns which have NAs
summary.Erie_dropNAs <- summary.Erie %>% 
  drop_na()
  
```

```{r, include = FALSE, message = FALSE, warning = FALSE}
#recode
summary.Erie_dropNAs$Variable <- dplyr:: recode(summary.Erie_dropNAs$Variable, 
                                         Dissolved.Microcystin.b5g.L = "Dissolved Microcystin (ug/L)",
                                         Extracted.Chlorophyll.a.b5g.L="Chlorophyll a (ug/L)",
                                         CTD.Dissolved.Oxygen.mg.L="Dissolved Oxygen (mg/L)",
                                         Nitrate..Nitrite.mg.N.L="Nitrate/Nitrite Concentration (mg/L)",
                                         Total.Dissolved.Phosphorus.b5g.P.L="Dissolved Phosphorus (ug/L",
                                         CTD.Temperature.b0C= "Temperature (deg C)")

```

```{r, warning = FALSE, meassage = FALSE, echo = FALSE}
#Create Table

summary.Erie_dropNAs %>%
gt() %>%

  cols_width(
    c(Variable)~ px(120),
    c(Obs)~ px(120),
    everything() ~ px(90))%>%

  tab_header(
    title = md("Summary Statistics: Lake Erie Harmful Algal Bloom Characteristics at Bottom and Surface Depths")) %>%
  
  fmt_passthrough (columns=c(Variable)) %>%
  fmt_number(columns = c(Obs), decimals = 0) %>%
  fmt_number(columns = c(Median), decimals = 2) %>%
  fmt_number(columns = c(Mean), decimals = 2) %>%
  fmt_number(columns = c(SD), decimals = 2) %>%
  fmt_number(columns = c(Max), decimals = 2) %>% 
  fmt_number(columns = c(Min), decimals = 2) %>% 
  
cols_label(
    Variable = "",
    Obs = "Oberservations",
    Median = "Median",
    Mean = "Mean",
    SD = "St Deviation",
    Max = "Max",
    Min="Min")
  
```


\newpage

# Analysis



## Question 1: How does Microcystis and microcystin concentration change over time?

Upon initial analysis of the data, there seems to be variation in bloom activity between different months and years. Chlorophyll a, as mentioned above, was highest in 2015, and seems to be declining overall from from 2012 to 2018, which had the lowest peaks in chlorophyll a of all the years (Fig. ###).  

```{r, include=FALSE}
Erie_data_subset <- Erie_Data_2012_2018 %>%
  filter(Extracted.Chlorophyll.a.b5g.L < 6000) %>%
  group_by(Date) %>%
  summarize(mean_chla = mean(Extracted.Chlorophyll.a.b5g.L),
            mean_temp = mean(CTD.Temperature.b0C)) %>%
  mutate(month = month(Date))

Erie_data_subset2 <- rbind(Erie_data_subset[1:15, ],
                               NA,
                               Erie_data_subset[16:37, ], 
                               NA,
                               Erie_data_subset[38:58, ],
                               NA,
                               Erie_data_subset[59:81, ],
                               NA,
                               Erie_data_subset[82:100, ],
                               NA,
                               Erie_data_subset[101:122, ],
                               NA,
                               Erie_data_subset[123:140, ])
#View(Erie_data_subset2)

Date_2012 <- as.Date("12/31/2012", format = "%m/%d/%Y")
Date_2013 <- as.Date("12/31/2013", format = "%m/%d/%Y")
Date_2014 <- as.Date("12/31/2014", format = "%m/%d/%Y")
Date_2015 <- as.Date("12/31/2015", format = "%m/%d/%Y")
Date_2016 <- as.Date("12/31/2016", format = "%m/%d/%Y")
Date_2017 <- as.Date("12/31/2017", format = "%m/%d/%Y")

Erie_data_subset2[16, 1] <- Date_2012
Erie_data_subset2[39, 1] <- Date_2013
Erie_data_subset2[61, 1] <- Date_2014
Erie_data_subset2[85, 1] <- Date_2015
Erie_data_subset2[105, 1] <- Date_2016
Erie_data_subset2[128, 1] <- Date_2017

#View(Erie_data_subset2)

```

```{r, warning=FALSE, message=FALSE, echo=FALSE}
#Plotting chlorophyll a from 2012 to 2018
chla_plot <- ggplot() +
  geom_line(data=Erie_data_subset2, aes(x=Date, y=mean_chla),
            na.rm=T) +
  ylab(expression("Chlorophyll a" ~ "("*mu*"g/L)")) +
  xlab("Year")

chla_plot

```
***Figure ###. Chlorophyll a concentrations from 2012-2018***

Temperature, by contrast, appears to have remained relatively constant between 2012 and 2018 (Fig. ##). 

```{r, warning=FALSE, message=FALSE, echo=FALSE}
#Plotting temperature from 2012 to 2018
temp_plot <- ggplot() +
  geom_line(data=Erie_data_subset2, aes(x=Date, y=mean_temp),
            na.rm=T) +
  ylab(expression("Temperature" ~ "("*degree*C*")")) +
  xlab("Year")

temp_plot

```
***Figure ###. Water temperature from 2012-2018***

It was difficult to visualize annual trends in microcystin, nitrate and nitrite, and total dissolved phosphorus concentrations since data collection is sporadic throughout the seven-year period.

We averaged chlorophyll a by month for each year to examine monthly trends. The monthly trends in chlorophyll a were similar across 2014, 2018, and 2022, with chlorophyll a peaking in August. This peak was greater in 2022 than in 2014 and 2018 (Fig. #).

```{r, include=FALSE}
#Create subset of data
Erie_data_subset <- data_frame(Erie_Data_2012_2018$Date,
                               Erie_Data_2012_2018$CTD.Temperature.b0C, 
                               Erie_Data_2012_2018$CTD.Dissolved.Oxygen.mg.L,
                               Erie_Data_2012_2018$Dissolved.Microcystin.b5g.L,
                               Erie_Data_2012_2018$Extracted.Chlorophyll.a.b5g.L, 
 
Erie_Data_2012_2018$Year)

colnames(Erie_data_subset) <- c("Date", "Temp", "DO", "MC", "Chla", "Year")
#View(Erie_data_subset)

#Group by month and year
Erie_data_month_means <- Erie_data_subset %>%
    mutate(Month = month(Date)) %>%
    group_by(Year, Month) %>%
    summarize(mean_chla = mean(Chla, na.rm=T),
              mean_temp = mean(Temp, na.rm=T),
              mean_mc = mean(MC, na.rm=T), 
              mean_do = mean(DO, na.rm=T))

#Group by month: 2022 data
class(Erie_Data_2022$Date)
Erie_Data_2022$Date <- as.Date(Erie_Data_2022$Date, format="%m/%d/%Y")

Erie_data_2022_means <- Erie_Data_2022 %>%
  mutate(Month = month(Date)) %>%
  mutate(Year = year(Date)) %>%
  group_by(Year, Month) %>%
  summarize(mean_chla = mean(Chla, na.rm=T),
            mean_temp = mean(Temp, na.rm=T),
            mean_mc = mean(MC, na.rm=T),
            mean_do = mean(DO, na.rm=T))

#Combine subsets
Erie_data_month_2012_2022 <- rbind(Erie_data_month_means, Erie_data_2022_means)
#View(Erie_data_month_2012_2022)

```

```{r, warning=FALSE, message=FALSE, echo=FALSE}
#Plot monthly data
chla_plot_monthly <- ggplot() +
  geom_line(data=filter(Erie_data_month_2012_2022, Year=="2014" | Year=="2018" | Year=="2022"), aes(x=Month, y=mean_chla, color=as.factor(Year)), linewidth=0.75) +
  scale_color_manual(values=c("#F8766D", "#00BFC4", "#7CAE00")) +
  xlab("Month") + 
  ylab(expression("Chlorophyll a" ~ "("*mu*"g/L)")) +
  ggtitle("Monthly Trends in Chlorophyll a by Year") +
  labs(color="Year")

chla_plot_monthly

```
***Figure ###. Monthly trends in chlorophyll a in 2014, 2018, and 2022.***

It was difficult to visualize annual and monthly trends in microcystin, nitrate and nitrite, and total dissolved phosphorus concentrations since data collection is sporadic throughout the seven-year period.

To assess whether differences in chlorophyll a and temperature between months and years were statistically significant, we conducted two-way ANOVA tests.

[ANOVAs here]

\newpage

## Question 2: Which variables (i.e., temperature, nitrate/nitrite, and/or phosphorus) contribute to microcystin production, Microcystis population growth, and dissolved oxygen?

**Generalized linear modeling**
To determine the variables driving chlorophyll a and microcystin concentration in Western Lake Erie, we conducted generalized linear modelling, taking into account temperature, nitrate and nitrite concentration, and total dissolved phosphorus concentration. We analyzed these trends for the entire 2012-2018 dataset, and for the years 2014 and 2018 specifically. 

The linear model results demonstrate a statistically significant negative relationship between microcystin concentration and temperature for the total dataset, but a significant positive relationship for data collected in 2014. Microcystin concentrations in 2018 were negatiely associated with nitrate and nitrite concentration (Table #). 

In 2018, chlorophyll a was positively associated with both water temperature and nitrate and nitrite concentration. However, in 2014 and across the total dataset, chlorophyll a was not significantly related to any of our explanatory variables (Table #). 

```{r, include=FALSE}
#Total dataset linear model results
#Chlorophyll a
Erie_chla_temp_lm <- lm(data=Erie_Data_2012_2018, Extracted.Chlorophyll.a.b5g.L ~ CTD.Temperature.b0C)
summary(Erie_chla_temp_lm)

Erie_chla_NO23_lm <- lm(data=Erie_Data_2012_2018, Extracted.Chlorophyll.a.b5g.L ~ Nitrate..Nitrite.mg.N.L)
summary(Erie_chla_NO23_lm)

Erie_chla_TDP_lm <- lm(data=Erie_Data_2012_2018, Extracted.Chlorophyll.a.b5g.L ~ Total.Dissolved.Phosphorus.b5g.P.L)
summary(Erie_chla_TDP_lm)

#Microcystin
Erie_MC_temp_lm <- lm(data=Erie_Data_2012_2018, Dissolved.Microcystin.b5g.L ~ CTD.Temperature.b0C)
summary(Erie_MC_temp_lm)

Erie_MC_NO23_lm <- lm(data=Erie_Data_2012_2018, Dissolved.Microcystin.b5g.L ~ Nitrate..Nitrite.mg.N.L)
summary(Erie_MC_NO23_lm)

Erie_MC_TDP_lm <- lm(data=Erie_Data_2012_2018, Dissolved.Microcystin.b5g.L ~ Total.Dissolved.Phosphorus.b5g.P.L)
summary(Erie_MC_TDP_lm)

```

```{r, include=FALSE}
#2014 linear model results
#Chlorophyll a
Erie_chla_temp_2014_lm <- lm(data=filter(Erie_Data_2012_2018, Year=="2014"), Extracted.Chlorophyll.a.b5g.L ~ CTD.Temperature.b0C)
summary(Erie_chla_temp_2014_lm)

Erie_chla_NO23_2014_lm <- lm(data=filter(Erie_Data_2012_2018, Year=="2014"), Extracted.Chlorophyll.a.b5g.L ~ Nitrate..Nitrite.mg.N.L)
summary(Erie_chla_NO23_2014_lm)

Erie_chla_TDP_2014_lm <- lm(data=filter(Erie_Data_2012_2018, Year=="2014"), Extracted.Chlorophyll.a.b5g.L ~ Total.Dissolved.Phosphorus.b5g.P.L)
summary(Erie_chla_TDP_2014_lm)

#Microcystin
Erie_MC_temp_2014_lm <- lm(data=filter(Erie_Data_2012_2018, Year=="2014"), Dissolved.Microcystin.b5g.L ~ CTD.Temperature.b0C)
summary(Erie_MC_temp_2014_lm)

Erie_MC_NO23_2014_lm <- lm(data=filter(Erie_Data_2012_2018, Year=="2014"), Dissolved.Microcystin.b5g.L ~ Nitrate..Nitrite.mg.N.L)
summary(Erie_MC_NO23_2014_lm)

Erie_MC_TDP_2014_lm <- lm(data=filter(Erie_Data_2012_2018, Year=="2014"), Dissolved.Microcystin.b5g.L ~ Total.Dissolved.Phosphorus.b5g.P.L)
summary(Erie_MC_TDP_2014_lm)

#2018 linear model results
#Chlorophyll a
Erie_chla_temp_2018_lm <- lm(data=filter(Erie_Data_2012_2018, Year=="2018"), Extracted.Chlorophyll.a.b5g.L ~ CTD.Temperature.b0C)
summary(Erie_chla_temp_2018_lm)

Erie_chla_NO23_2018_lm <- lm(data=filter(Erie_Data_2012_2018, Year=="2018"), Extracted.Chlorophyll.a.b5g.L ~ Nitrate..Nitrite.mg.N.L)
summary(Erie_chla_NO23_2018_lm)

Erie_chla_TDP_2018_lm <- lm(data=filter(Erie_Data_2012_2018, Year=="2018"), Extracted.Chlorophyll.a.b5g.L ~ Total.Dissolved.Phosphorus.b5g.P.L)
summary(Erie_chla_TDP_2018_lm)

#Microcystin
Erie_MC_temp_2018_lm <- lm(data=filter(Erie_Data_2012_2018, Year=="2018"), Dissolved.Microcystin.b5g.L ~ CTD.Temperature.b0C)
summary(Erie_MC_temp_2018_lm)

Erie_MC_NO23_2018_lm <- lm(data=filter(Erie_Data_2012_2018, Year=="2018"), Dissolved.Microcystin.b5g.L ~ Nitrate..Nitrite.mg.N.L)
summary(Erie_MC_NO23_2018_lm)

Erie_MC_TDP_2018_lm <- lm(data=filter(Erie_Data_2012_2018, Year=="2018"), Dissolved.Microcystin.b5g.L ~ Total.Dissolved.Phosphorus.b5g.P.L)
summary(Erie_MC_TDP_2018_lm)

```

```{r, include=FALSE}
#Create dataframes from linear model summaries
chla_temp_df <- tidy(Erie_chla_temp_lm)

chla_no23_df <- tidy(Erie_chla_NO23_lm)

chla_tdp_df <- tidy(Erie_chla_TDP_lm)

mc_temp_df <- tidy(Erie_MC_temp_lm)

mc_no23_df <- tidy(Erie_MC_NO23_lm)

mc_tdp_df <- tidy(Erie_MC_TDP_lm)


chla_temp_2014_df <- tidy(Erie_chla_temp_2014_lm)

chla_no23_2014_df <- tidy(Erie_chla_NO23_2014_lm)

chla_tdp_2014_df <- tidy(Erie_chla_TDP_2014_lm)

mc_temp_2014_df <- tidy(Erie_MC_temp_2014_lm)

mc_no23_2014_df <- tidy(Erie_MC_NO23_2014_lm)

mc_tdp_2014_df <- tidy(Erie_MC_TDP_2014_lm)


chla_temp_2018_df <- tidy(Erie_chla_temp_2018_lm)

chla_no23_2018_df <- tidy(Erie_chla_NO23_2018_lm)

chla_tdp_2018_df <- tidy(Erie_chla_TDP_2018_lm)

mc_temp_2018_df <- tidy(Erie_MC_temp_2018_lm)

mc_no23_2018_df <- tidy(Erie_MC_NO23_2018_lm)

mc_tdp_2018_df <- tidy(Erie_MC_TDP_2018_lm)

#Merging dataframes 
linear_model_results <- rbind(chla_temp_df, chla_no23_df, chla_tdp_df,
                              mc_temp_df, mc_no23_df, mc_tdp_df,
                              chla_temp_2014_df, chla_no23_2014_df, 
                              chla_tdp_2014_df,
                              mc_temp_2014_df, mc_no23_2014_df, 
                              mc_tdp_2014_df,
                              chla_temp_2018_df, chla_no23_2018_df, 
                              chla_tdp_2018_df,
                              mc_temp_2018_df, mc_no23_2018_df, 
                              mc_tdp_2018_df)

linear_model_results <- data_frame(linear_model_results$term, linear_model_results$estimate, linear_model_results$std.error, linear_model_results$p.value)

colnames(linear_model_results) = c("Term", "Coefficient", "Standard_error", "p_value")

linear_model_results$year = c("Total", "Total", "Total","Total", "Total", "Total", "Total", "Total", "Total","Total", "Total", "Total", "2014", "2014", "2014","2014", "2014", "2014", "2014", "2014", "2014","2014", "2014", "2014", "2018", "2018", "2018","2018", "2018", "2018", "2018", "2018", "2018","2018", "2018", "2018")

linear_model_results$response_variable = c("Chlorophyll a", "Chlorophyll a", "Chlorophyll a", "Chlorophyll a", "Chlorophyll a", "Chlorophyll a", "Microcystin", "Microcystin", "Microcystin", "Microcystin", "Microcystin", "Microcystin", "Chlorophyll a", "Chlorophyll a", "Chlorophyll a", "Chlorophyll a", "Chlorophyll a", "Chlorophyll a", "Microcystin", "Microcystin", "Microcystin", "Microcystin", "Microcystin", "Microcystin", "Chlorophyll a", "Chlorophyll a", "Chlorophyll a", "Chlorophyll a", "Chlorophyll a", "Chlorophyll a", "Microcystin", "Microcystin", "Microcystin", "Microcystin", "Microcystin", "Microcystin")

#Changing variable names
linear_model_results$Term[linear_model_results$Term == "CTD.Temperature.b0C"] <- "Temperature"

linear_model_results$Term[linear_model_results$Term == "Nitrate..Nitrite.mg.N.L"] <- "Nitrate + Nitrite"

linear_model_results$Term[linear_model_results$Term == "Total.Dissolved.Phosphorus.b5g.P.L"] <- "Total Dissolved Phosphorus"

linear_model_results <- linear_model_results %>%
  group_by(year, response_variable) %>%
  filter(!Term=="(Intercept)")

#View(linear_model_results)

```

***Table ###. Linear model results of relationships between response variables: chlorophyll a and microcystin, and explanatory variables: temperature, nitrate and nitrite, and total dissolved phosphorus concentration***
```{r, error=FALSE, message=FALSE, echo=FALSE}
#Create table of linear model results
linear_model_results %>%
gt() %>%
 
  cols_width(
    c(Term)~ px(120),
    c(Coefficient)~ px(120),
    c(Standard_error)~ px(120),
    c(p_value)~ px(120),
    everything() ~ px(90))%>%

  tab_header(
    title = md("2012-2018 chlorophyll a and microcystin linear model results")) %>%

  fmt_passthrough (columns=c(Term)) %>%
  fmt_number(columns = c(Coefficient), decimals = 5) %>%
  fmt_number(columns = c(Standard_error), decimals = 5) %>%
  fmt_number(columns = c(p_value), decimals = 3) %>%
  
   tab_style(
    style = list(
      cell_text(weight = "bold")),
    locations = cells_body(
      columns = p_value,
      rows = p_value <= 0.05)) %>%
  
  sub_small_vals() %>%

cols_label(
    Term = "Explanatory variable",
    Coefficient = "Coefficient",
    Standard_error = "Standard error",
    p_value = "p-value")

```

\newpage
The plots below illustrate the linear model results. As shown in Fig. #, the nonsignificant positive relationship between temperature and chlorophyll a from 2012 to 2018 is fairly weak and may be influenced by extremely high observations, like those in 2015. The lack of significant relationship between chlorophyll a and temperature in 2014 which was present in 2018 may be due to the temperature range. There seems to be less variety in temperature measurements in 2018 than 2014 (Fig. #).

```{r, warning=FALSE, message=FALSE, echo=FALSE}
#Plotting linear model results 
#Total dataset: chlorophyll a --> maybe take out?
chla_plot <- ggplot(data=Erie_Data_2012_2018, aes(x=CTD.Temperature.b0C, y=Extracted.Chlorophyll.a.b5g.L)) +
  geom_point() +
  geom_smooth(method=lm, se=F, color="#009E73") +
  ylim(0, 500) +
  xlab(expression("Temperature" ~ "("*degree*C*")")) + 
  ylab(expression("Chlorophyll a" ~ "("*mu*"g/L)")) +
  ggtitle("Relationship between Chlorophyll a and Temperature")
chla_plot

```
***Figure #. Line plot showing relationship between chlorophyll a (y axis) and water temperature (x axis) between 2012 and 2018. Observations above 500 were removed to improve data visualization***

```{r, warning=FALSE, message=FALSE, echo=FALSE}
#2014 and 2018: chloropyll a
chla_plot2 <- ggplot(data=filter(Erie_Data_2012_2018, Year=="2014"| Year=="2018"), aes(x=CTD.Temperature.b0C, y=Extracted.Chlorophyll.a.b5g.L, color=as.factor(Year))) +
  geom_point(color="black") +
  geom_smooth(method=lm, se=F) +
  ylim(0, 130) +
  xlab(expression("Temperature" ~ "("*degree*C*")")) + 
  ylab(expression("Chlorophyll a" ~ "("*mu*"g/L)")) +
  labs(color="Year") +
  ggtitle("Relationship between Chlorophyll a and Temperature")
chla_plot2

```
***Figure #. Line plot showing relationship between chlorophyll a (y axis) and water temperature (x axis) in 2014 (red line) and 2018 (blue line).***

While microcystin concentrations between 2012 and 2018 show a negative trend with temperature overall (Fig. #), there was a significant, positive association between microcystin and temperature in 2014 (Fig. #). As discussed above, microcystin collection was sporadic throughout the study period, and the low sample size may affect our ability to detect relationships between microcystin concentration and water quality variables.

```{r, warning=FALSE, message=FALSE, echo=FALSE}
#Plotting linear model results
#Total dataset: microcystin
mc_plot <- ggplot(data=Erie_Data_2012_2018, aes(x=CTD.Temperature.b0C, y=Dissolved.Microcystin.b5g.L)) +
  geom_point() +
  geom_smooth(method=lm, se=F, color="#009E73") +
  ylim(0, 2) +
  #scale_y_continuous(trans='log10') +
  xlab(expression("Temperature" ~ "("*degree*C*")")) + 
  ylab(expression("Microcystin" ~ "("*mu*"g/L)")) +
  ggtitle("Relationship between Microcystin and Temperature")
mc_plot

```
***Figure #. Line plot showing relationship between microcystin (y axis) and water temperature (x axis) between 2012 and 2018.***

```{r, warning=FALSE, message=FALSE, echo=FALSE}
#2014 and 2018: microcystin
mc_plot2 <- ggplot(data=filter(Erie_Data_2012_2018, Year=="2014"| Year=="2018"), aes(x=CTD.Temperature.b0C, y=Dissolved.Microcystin.b5g.L, color=as.factor(Year))) +
  geom_point(color="black") +
  geom_smooth(method=lm, se=F) +
  ylim(0, 2) +
  xlab(expression("Temperature" ~ "("*degree*C*")")) + 
  ylab(expression("Microcystin" ~ "("*mu*"g/L)")) +
  labs(color="Year") +
  ggtitle("Relationship between Microcystin and Temperature")
mc_plot2

```
***Figure #. Line plot showing relationship between microcystin (y axis) and water temperature (x axis) in 2014 (red line) and 2018 (blue line).***

In 2018, chlorophyll a was positively associated with nitrate and nitrite concentrations, but microcystin was negatively associated with nitrate and nitrite (Fig. #). This result was unexpected, as we hypothesized that both chlorophyll a and microcystin would respond similarly to water quality variables like nutrient concentration. 

```{r, warning=FALSE, message=FALSE, echo=FALSE}
#Plotting linear model results
#2018: nitrate and nitrite
#Chlorophyll a
chla_2018_plot <- ggplot(data=filter(Erie_Data_2012_2018, Year=="2018"), aes(x=Nitrate..Nitrite.mg.N.L, y=Extracted.Chlorophyll.a.b5g.L)) +
  geom_point() +
  geom_smooth(method=lm, se=F, color="#00BFC4") +
  ylim(0, 100) +
  xlim(0, 9) +
  xlab("Nitrate + Nitrite (mg N/L)") + 
  ylab(expression("Chlorophyll a" ~ "("*mu*"g/L)")) +
  ggtitle("Relationship between Chlorophyll a and Nitrate + Nitrite")
chla_2018_plot

#Microcystin
mc_2018_plot <- ggplot(data=filter(Erie_Data_2012_2018, Year=="2018"), aes(x=Nitrate..Nitrite.mg.N.L, y=Dissolved.Microcystin.b5g.L)) +
  geom_point() +
  geom_smooth(method=lm, se=F, color="#00BFC4") +
  ylim(0, 0.65) +
  xlim(0, 9) +
  xlab("Nitrate + Nitrite (mg N/L)") + 
  ylab(expression("Microcystin" ~ "("*mu*"g/L)")) +
  ggtitle("Relationship between Microcystin and Nitrate + Nitrite")
mc_2018_plot

```
***Figure #. Line plot showing relationship between microcystin (y axis) and nitrate + nitrite concentration (x axis) in 2018.***


We further used linear modeling to assess the relationships between chlorophyll a and microcystin and water temperature in 2022. In line with our hypotheses, both chlorophyll a and microcystin are positively related to water temperature (Table #).

```{r, include=FALSE}
#Linear model: chlorophyll a and temperature
Erie_chla_2022_lm <- lm(data=Erie_Data_2022, Chla ~ Temp)
summary(Erie_chla_2022_lm)

#Linear model: microcystin and temperature
Erie_mc_2022_lm <- lm(data=Erie_Data_2022, MC ~ Temp)
summary(Erie_mc_2022_lm)

```

```{r, include=FALSE}
#Create dataframe of linear modeling results
Erie_chla_2022_lm_df <- tidy(Erie_chla_2022_lm)

Erie_mc_2022_lm_df <- tidy(Erie_mc_2022_lm)

linear_model_results_2022 <- rbind(Erie_chla_2022_lm_df,
                                   Erie_mc_2022_lm_df)

linear_model_results_2022 <- linear_model_results_2022 %>%
  filter(!term=="(Intercept)")

linear_model_results_2022 <- data_frame( linear_model_results_2022$estimate, linear_model_results_2022$std.error, linear_model_results_2022$p.value)

colnames(linear_model_results_2022) = c("Coefficient", "Standard_error", "p_value")

linear_model_results_2022$response_variable = c("Chlorophyll a", "Microcystin")

linear_model_results_2022 <- linear_model_results_2022[ , c(4, 1, 2, 3)]

#View(linear_model_results_2022)

```

***Table #. Linear model results of relationships between response variables: chlorophyll a and microcystin, and temperature in 2022***
```{r, , warning=FALSE, message=FALSE, echo=FALSE}
#Create table of linear modeling results
linear_model_results_2022 %>%
gt() %>%
 
  cols_width(
    c(response_variable)~ px(120),
    c(Coefficient)~ px(120),
    c(Standard_error)~ px(120),
    c(p_value)~ px(120),
    everything() ~ px(90))%>%

  tab_header(
    title = md("2022 temperature linear model results")) %>%

  fmt_passthrough (columns=c(response_variable)) %>%
  fmt_number(columns = c(Coefficient), decimals = 5) %>%
  fmt_number(columns = c(Standard_error), decimals = 5) %>%
  fmt_number(columns = c(p_value), decimals = 3) %>%
  
   tab_style(
    style = list(
      cell_text(weight = "bold")),
    locations = cells_body(
      columns = p_value,
      rows = p_value <= 0.05)) %>%
  
  sub_small_vals() %>%

cols_label(
    response_variable = "Response variable",
    Coefficient = "Coefficient",
    Standard_error = "Standard error",
    p_value = "p-value")

```



## Question 3: Are there differences between toxin and algae concentrations at different lake depths?

Regarding the third question it is hypothesized that both microcystin, chlorophyll a, and dissolved oxygen concentrations will be greater at surface depths opposed to bottom depths in Lake Erie. These hypotheses are stated below:

***Hypotheses for microcystin concentration at surface vs. bottom depth***

Ho: Surface depth microcystin concentrations are <= bottom depth microcystin concentrations

HA: Surface depth microcystin concentrations are > bottom depth microcystin concentrations

***Hypotheses for chlorophyll a concentration at surface vs. bottom depth***

Ho: Surface depth Chlorophyll a concentrations are <= bottom depth microcystin concentrations

HA: Surface depth Chlorophyll a concentrations are > bottom depth microcystin concentrations

***Hypotheses for dissolved oxygen concentration at surface vs. bottom depth***

Ho: Surface depth dissolved oxygen concentrations are <= bottom depth microcystin concentrations

HA: Surface depth dissolved oxygen concentrations are > bottom depth microcystin concentrations

**Assessing distribution of data**
Before assessing these claims, a Shapiro-Wilk test for normality was conducted for the microcystin, chlorophyll a, and dissolved oxygen data. The test rejected the null hypothesis which states that the data are normally distributed in all cases as shown in Table ###. In result, the Wilcoxon Rank Sum test was employed to evaluate the means between surface and bottom layers for each variable. These results can be found in Table ###. 

```{r, include=FALSE}
#test for normality of data
shapiro.microcystin <- shapiro.test(bottom_surface_Erie$Dissolved.Microcystin.b5g.L) #the data is not shown to be normal; reject the null hypothesis of normality in this case

shapiro.Chla <- shapiro.test(bottom_surface_Erie$Extracted.Chlorophyll.a.b5g.L) #the data is not shown to be normal; reject the null hypothesis of normality in this case

shapiro.DO <- shapiro.test(bottom_surface_Erie$CTD.Dissolved.Oxygen.mg.L)#the data is not shown to be normal; reject the null hypothesis of normality in this case
```

```{r, include=FALSE}
#Shapiro-wilk test table formatting
shapiro.1 <- rbind(shapiro.microcystin, shapiro.Chla, shapiro.DO)
shapiro<- as.data.frame(shapiro.1)

shapiro <- shapiro %>% 
  select(!method)

colnames_data <- c("Dissolved Microcystin", "Chlorophyll a", "Dissolved Oxygen")

shapiro$data.name <- 
  replace(shapiro$data.name,values = colnames_data)

shapiro.10 <- shapiro %>% 
  mutate_at(vars(1,2),  funs(gsub("[c()W=]", "", .)))

shapiro.10_num <- shapiro.10 %>% 
  mutate_if(is.character, as.numeric)

shapiro.10_round <- shapiro.10_num %>% 
  mutate_at(vars(1,2),  funs(round(.,digits = 3)))

colnames(shapiro.10_round) <- c('Statistic','P-Value','Data')

shapiro_reorder <- shapiro.10_round[, c(3, 1, 2)]

```

***Table ###. Shapiro-Wilk Test for Normality Results***

```{r, warning = FALSE, meassage = FALSE, echo = FALSE}
gt(shapiro_reorder) %>% 
  tab_header(
    title = "Shapiro-Wilk Test Results")
```

```{r, include = FALSE}
#Drop NAs

bottom_surface_Erie_dropNA_microcystin <- bottom_surface_Erie %>% 
  drop_na(Dissolved.Microcystin.b5g.L)

bottom_surface_Erie_dropNA_CHLa <- bottom_surface_Erie %>% 
  drop_na(Extracted.Chlorophyll.a.b5g.L)

bottom_surface_Erie_dropNA_DO <- bottom_surface_Erie %>% 
  drop_na(CTD.Dissolved.Oxygen.mg.L)
```

```{r, include = FALSE}
#Microcystin: surface vs bottom

wilcox_microcystin <- wilcox.test(Dissolved.Microcystin.b5g.L ~ Sample.Depth.category, data = bottom_surface_Erie_dropNA_microcystin, alternative = "less") #Due to nonnormality of the data, a rank sum test was applied

wilcox_microcystin

#CHLa: surface vs bottom

wilcox_chla <- wilcox.test(Extracted.Chlorophyll.a.b5g.L ~ Sample.Depth.category, data = bottom_surface_Erie_dropNA_CHLa, alternative = "less") #Due to nonnormality of the data, a rank sum test was applied

wilcox_chla

wilcox_DO <- wilcox.test(CTD.Dissolved.Oxygen.mg.L ~ Sample.Depth.category, data = bottom_surface_Erie_dropNA_DO, alternative = "less") #Due to nonnormality of the data, a rank sum test was applied

wilcox_DO
```

```{r, include=FALSE}
#Wilcoxin Rank Sum test table formatting
wilcox.1 <- rbind(wilcox_microcystin, wilcox_chla, wilcox_DO)
wilcox<- as.data.frame(wilcox.1)

wilcox <- wilcox %>% 
  select(statistic, p.value, data.name)

colnames_data_w <- c("Dissolved Microcystin", "Chlorophyll a", "Dissolved Oxygen")

wilcox$data.name <- 
  replace(wilcox$data.name,values = colnames_data_w)

wilcox.10 <- wilcox %>% 
  mutate_at(vars(1,2),  funs(gsub("[c()W=]", "", .)))

wilcox.10_num <- wilcox.10 %>% 
  mutate_if(is.character, as.numeric)

wilcox.10_round <- wilcox.10_num %>% 
  mutate_at(vars(1,2),  funs(round(.,digits = 3)))

colnames(wilcox.10_round) <- c('Statistic','P-Value','Data')

wilcox_reorder <- wilcox.10_round[, c(3, 1, 2)]

```

***Table ###. Wilcoxin Rank Sum Test for Bottom versus Surface Depths***
```{r, warning = FALSE, meassage = FALSE, echo = FALSE}
gt(wilcox_reorder) %>% 
  tab_header(
    title = "Wilcoxin Rank Sum Test Results")
```

When looking at microcystin concentrations at surface and bottom depths, the Wilcoxin Rank Sum test showed a nonsignificant result and thus a failure to reject the null hypothesis which states that microcystin concentrations at surface depth are less than or equal to microcystin concentrations at bottom depth (W = 24800, p-value = 0.1753). Figure ### shows these data. When looking at chlorophyll a concentrations at surface and bottom depths, the Wilcoxin Rank Sum test showed a significant result and a rejection of the null hypothesis which states that chlorophyll a concentrations at surface depth are less than or equal to chlorophyll a concentrations at bottom depth (W = 78212, p-value = 0.0008634). Figure ### shows these data. Lastly, when looking at dissolved oxygen concentrations, the Wilcoxin Rank Sum test showed a significant result and a rejection of the null hypothesis which states that dissolved oxygen concentrations at surface depth are less than or equal to dissolved concentrations at bottom depth (W = 51924, p-value < 2.2e-16). Figure ### shows these data.

```{r, warning = FALSE, meassage = FALSE, echo = FALSE}
#Microcystin boxplot
Microcystin_SB_plot <- ggplot(bottom_surface_Erie_dropNA_microcystin)+ 
  geom_boxplot(aes(x = as.factor(Year), y=Dissolved.Microcystin.b5g.L, fill = Sample.Depth.category))+
  ylim(0,0.75)+
  labs(y = "Microcystin Concentration (ug/L)", x = "Sample Depth Category", title = "Microcystin Concentration at Surface and Bottom Depths")

print(Microcystin_SB_plot)
```
***Figure ###. Microcystin Concentrations at Surface and Bottom Depth for years 2015 through 2018***

```{r, warning = FALSE, meassage = FALSE, echo = FALSE}
#CHLa boxplot: surface vs bottom
CHLa_SB_plot <- ggplot(bottom_surface_Erie_dropNA_CHLa)+ 
  geom_boxplot(aes(x = as.factor(Year), y=Extracted.Chlorophyll.a.b5g.L, fill = Sample.Depth.category))+
  ylim(0,100)+
  labs(y = "Extracted Chlorophyll a Concentration (ug/L)", x = "Sample Depth Category", title = "Extracted Chlorophyll a Concentration at Surface and Bottom Depths")

print(CHLa_SB_plot)
```

***Figure ###. Chlorophyll a Concentrations at Surface and Bottom Depth for years 2015 through 2018***

```{r, warning = FALSE, meassage = FALSE, echo = FALSE}
#DO boxplot: surface vs bottom
DO_SB_plot <- ggplot(bottom_surface_Erie_dropNA_DO)+ 
  geom_boxplot(aes(x = as.factor(Year), y=CTD.Dissolved.Oxygen.mg.L, fill = Sample.Depth.category))+
  labs(y = "Dissolved Oxygen Concentration (mg/L)", x = "Sample Depth Category", title = "Dissolved Oxygen Concentration at Surface and Bottom Depths")

print(DO_SB_plot)
```

***Figure ###. Dissolved Oxygen Concentrations at Surface and Bottom Depth for years 2015 through 2018***


\newpage
# Summary and Conclusions


\newpage

# References

Centers for Disease Control and Prevention (CDC; 2022). Causes and Ecosystem Impacts. https://www.cdc.gov/habs/environment.html#:~:text=Levels%20of%20nutrients%20such
%20as%20phosphorus%20and%20nitrogen%20increase%20in,Sewage%20from%20people%20and
%20animals

Dean, S. (2022). Harmful algal blooms in Lake Erie expected to be smaller this year, says NOAA. https://www.michigan.gov/egle/newsroom/mi-environment/2022/08/11/harmful-algal-blooms-in-lake-erie-expected-to-be-smaller-this-year-says-noaa#:~:text=The%20Lake%20Erie%20HABs%20consist,treatment%20of%20their%20source%20water.
Dierkes, C. (2014). Algae Through the Ages. https://ohioseagrant.osu.edu/news/2014/17gmc/algae-through-the-ages

Lake Erie Foundation (LEF; n.d.) Harmful Algal Blooms. https://lakeeriefoundation.org/issues/nutrientsharmful-algae/#:~:text=Harmful%20algal%20blooms%20(HABs)%20are,and%20the%20resulting%20algal%20blooms.

National Oceanic and Atmospheric Administration (NOAA; 2016). What is a harmful algal bloom? https://www.noaa.gov/what-is-harmful-algal-bloom

National Science Foundation (NSF; 2019). Lake Erie's toxic algae blooms: Why is the water turning green? https://beta.nsf.gov/news/lake-eries-toxic-algae-blooms-why-water-turning#:~:text=Since%20the%20late%201990s%2C%20Lake,warnings%20to%20avoid%20water%20contact. 

United States Environmental Protection Agency (U.S. EPA; 2022a). Harmful Algal Blooms. https://www.epa.gov/nutrientpollution/harmful-algal-blooms

United States Environmental Protection Agency (U.S. EPA; 2022b). Lake Erie. https://www.epa.gov/greatlakes/lake-erie
U.S. National Office for HABs. Recent Trends: National Changes. (n.d.) https://hab.whoi.edu/maps/regions-us-distribution/regions-us-recent-trends/
```

