---
output:
  pdf_document: default
  html_document: default
---
\newpage
\tableofcontents 
\newpage
\listoftables 
\newpage
\listoffigures 
\newpage

```{r setup, include=FALSE}
# Set your working directory
getwd()

# Load your packages
library(tidyverse)
#install.packages("lubridate")
library(lubridate)
#install.packages("zoo")
library(zoo)
#install.packages("trend")
library(trend)
library(ggplot2)
#install.packages("Kendall")
library(Kendall) 
#install.packages("gt")
library(gt)

# Set your ggplot theme
ggplot_theme <- theme_linedraw(base_size = 12) +
  theme(panel.grid=element_line(colour="gray"),
        axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(ggplot_theme)

# Load your datasets
Erie_Data_2012_2018 <- 
  read.csv("./Data/Raw/lake_erie_habs_field_sampling_results_2012_2018_v2.csv",
           check.names = F)

#Erie_Data_2012_2018 <-   read.csv("./Data/Processed/Erie_2012_2018_processed.csv",
          # check.names = F)
```

```{r, include = FALSE}
#Clean up column names
colnames(Erie_Data_2012_2018) <- gsub(" ", ".", colnames(Erie_Data_2012_2018))
colnames(Erie_Data_2012_2018) <- gsub("/", ".", colnames(Erie_Data_2012_2018))
colnames(Erie_Data_2012_2018) <- gsub("[//+]", "", colnames(Erie_Data_2012_2018))
colnames(Erie_Data_2012_2018) <- gsub("[()]", "", colnames(Erie_Data_2012_2018))
colnames(Erie_Data_2012_2018) <- gsub("[<>]", "", colnames(Erie_Data_2012_2018))

colnames(Erie_Data_2012_2018)

colnames(Erie_Data_2012_2018)[19] <- "Photo.active.radiation"
```

```{r, include = FALSE}
#Create new data frame 
Erie_Data_2012_2018 <- data_frame(Erie_Data_2012_2018$Date,
                                  Erie_Data_2012_2018$Site,
                                  Erie_Data_2012_2018$Sample.Depth.category,
                                  Erie_Data_2012_2018$Dissolved.Organic.Carbon.mg.L,
                                  Erie_Data_2012_2018$Urea.b5g.N.L,
                                  Erie_Data_2012_2018$Nitrate..Nitrite.mg.N.L,
                                  Erie_Data_2012_2018$Ammonia.b5g.N.L,
                                  Erie_Data_2012_2018$Soluble.Reactive.Phosphorus.b5g.P.L,
                                  Erie_Data_2012_2018$Total.Dissolved.Phosphorus.b5g.P.L,
                                  Erie_Data_2012_2018$Total.Phosphorus.b5g.P.L,
                                  Erie_Data_2012_2018$Extracted.Chlorophyll.a.b5g.L,
                                  Erie_Data_2012_2018$Extracted.Phycocyanin.b5g.L,
                                  Erie_Data_2012_2018$Dissolved.Microcystin.b5g.L,
                                  Erie_Data_2012_2018$CTD.Dissolved.Oxygen.mg.L,
                                  Erie_Data_2012_2018$Photo.active.radiation,
                                  Erie_Data_2012_2018$Sample.Temperature.b0C,
                                  Erie_Data_2012_2018$CTD.Temperature.b0C)
colnames(Erie_Data_2012_2018) <- gsub("Erie_Data_2012_2018", "", colnames(Erie_Data_2012_2018))
colnames(Erie_Data_2012_2018) <- gsub("[//$]", "", colnames(Erie_Data_2012_2018))
```

```{r, include = FALSE}
#change character variables to numeric; change date column to date format
lapply(Erie_Data_2012_2018, class)
Erie_Data_2012_2018$Dissolved.Microcystin.b5g.L <- as.numeric(Erie_Data_2012_2018$Dissolved.Microcystin.b5g.L)
Erie_Data_2012_2018$Extracted.Phycocyanin.b5g.L <- as.numeric(Erie_Data_2012_2018$Extracted.Phycocyanin.b5g.L)
Erie_Data_2012_2018$Soluble.Reactive.Phosphorus.b5g.P.L <- as.numeric(Erie_Data_2012_2018$Soluble.Reactive.Phosphorus.b5g.P.L)
Erie_Data_2012_2018$Ammonia.b5g.N.L <- as.numeric(Erie_Data_2012_2018$Ammonia.b5g.N.L)
Erie_Data_2012_2018$Nitrate..Nitrite.mg.N.L <- as.numeric(Erie_Data_2012_2018$Nitrate..Nitrite.mg.N.L)
Erie_Data_2012_2018$Urea.b5g.N.L <- as.numeric(Erie_Data_2012_2018$Urea.b5g.N.L)
Erie_Data_2012_2018$Date <- as.Date(Erie_Data_2012_2018$Date, format = "%m/%d/%Y")
```

```{r, include=FALSE}
#GLMs with surface and bottom variables
bottom_surface_Erie <- Erie_Data_2012_2018 %>% 
  filter(Sample.Depth.category == "Bottom" | Sample.Depth.category == "Surface")

Microcystin.model_SB <- lm(data = bottom_surface_Erie, Dissolved.Microcystin.b5g.L ~ 
                        Sample.Depth.category)

summary(Microcystin.model_SB) #there is a significant relationship between bottom and surface depths, but not enough data in the bottom section to be added as a dummy variable in MLR. I will break out to models into Surface and remove other observations. 

Erie_Surface_2012_2018 <- Erie_Data_2012_2018 %>% 
  filter(Sample.Depth.category == "Surface")

#Surface Microcystin model (explanatory variables include temperature, phosphorus, and nitrogen)
Microcystin.model_Surface <- lm(data = Erie_Surface_2012_2018, Dissolved.Microcystin.b5g.L ~ CTD.Temperature.b0C+ Total.Dissolved.Phosphorus.b5g.P.L+Nitrate..Nitrite.mg.N.L)

summary(Microcystin.model_Surface)
```

# Rationale and Research Questions

Harmful algal blooms (HABs) are characterized by rapid algae population growth coupled with toxin production in an aquatic system (NOAA, 2016). In recent history HABs have been shown to be increasing in prevalence with documented observations throughout the U.S. and the world (U.S. National Office for HABs, n.d.). This increased prevalence of HABs poses a threat to both public and ecosystem health (U.S. EPA, 2022a). Generally, HABs are associated with an influx of environmental nutrients into an aquatic system; however, other regional or lake-specific variables such as temperature, physical chemical properties, and lake structure can influence the development of HABs (CDC, 2022). Although HABs have been largely stated to have increased through time, many of these variables are challenging to generalize across water bodies due to the complexity of algal bloom formation. It is for this reason that this report will focus on HAB development in one water body: Lake Erie.  

Lake Erie has been affected by seasonal HABs since the 1990s (NSF, 2019). Lake Erie is adjacent to multiple metropolitan areas with populations that exceed 50,000 people and provides drinking water for 12 million people (U.S. EPA, 2022b). In 2015 Lake Erie had a bloom which covered over 300 square miles, thus making it one of the largest algal blooms documented in recent history (LEF, n.d.). Algae and toxin production has public health implications for surrounding populations as exposure can occur through recreation and consumption of contaminated drinking water (Dierkes, 2014; U.S. EPA, 2022b). Algal blooms can also create anoxic conditions when algae undergo decomposition in the environment which also influences ecosystem health (CDC, 2022). Due to the historical and current public health and ecosystem prevalence, data collection efforts have been implemented within Lake Erie to aid with forecasting future bloom severity. These data in the western basin are evaluated in this report.

Lake Erie is the shallowest of the great lakes with an average depth of 19 meters (m); the western portion, which comprises approximately 20% of the lake, has an average depth of 7.4 m and a maximum depth of 19 m (U.S. EPA, 2022b). Western Lake Erie is expected to undergo stratification for a short period of time during the summer months, leading to a warmer surface layer (epilimnion) and cooler bottom layer (hypolimnion; U.S. EPA, 2022b). Nutrient inputs come from a variety of sources which include wastewater treatment plants and agriculture which can influence algal bloom production (Dean, 2022). Microcystis is the most common species and microcystin, a potent hepatotoxin, is the most common toxin documented in HABs in Lake Erie. Considering the characteristics of Western Lake Erie, this report evaluates the following questions:

  What are the drivers of harmful algal blooms in western Lake Erie?
  
  1. Which variables (i.e., temperature, nitrate/nitrite, and/or phosphorus) contribute to microcystin production, Microcystis population growth, and dissolved oxygen?
  
  2. How does Microcystis and microcystin concentration change over time?
  
  3. Are there differences between toxin and algae concentrations at different lake depths?  
  
Within the following analysis, temperature, nutrient concentrations (i.e., nitrogen and phosphorus), and lake depth are used to assess these trends. 


\newpage

# Dataset Information






\newpage

# Exploratory Analysis 

***Table 1. Summary Statistics: Harmful Algal Bloom and Lake Characteristics of Lake Erie at Bottom and Surface Depths***
```{r, include = FALSE, message = FALSE, warning = FALSE}
# Select Variables of Interest

Erie.select<-bottom_surface_Erie%>%
 dplyr::select(Sample.Depth.category, Dissolved.Microcystin.b5g.L, Extracted.Chlorophyll.a.b5g.L, CTD.Dissolved.Oxygen.mg.L, Nitrate..Nitrite.mg.N.L, Total.Dissolved.Phosphorus.b5g.P.L, CTD.Temperature.b0C)
head(Erie.select)

```

```{r, include = FALSE, message = FALSE, warning = FALSE}

Erie.select.longer<-pivot_longer(Erie.select, !Sample.Depth.category, names_to = "Variable", values_to = "value")

summary.Erie<-Erie.select.longer%>%
  dplyr::group_by(Sample.Depth.category, Variable) %>%
  dplyr::summarize(
    Obs=length(value)-sum(is.na(value)),
    Mean=mean(value, na.rm=TRUE),
    SD=sd(value, na.rm=TRUE),
    Median=median(value, na.rm=TRUE))
   
head(summary.Erie)

#remove columns which have NAs
summary.Erie_dropNAs <- summary.Erie %>% 
  drop_na()
  
```

```{r, include = FALSE, message = FALSE, warning = FALSE}
#recode
summary.Erie_dropNAs$Variable <- dplyr:: recode(summary.Erie_dropNAs$Variable, 
                                         Dissolved.Microcystin.b5g.L = "Dissolved Microcystin (ug/L)",
                                         Extracted.Chlorophyll.a.b5g.L="Chlorophyll a (ug/L)",
                                         CTD.Dissolved.Oxygen.mg.L="Dissolved Oxygen (mg/L)",
                                         Nitrate..Nitrite.mg.N.L="Nitrate/Nitrite Concentration (mg/L)",
                                         Total.Dissolved.Phosphorus.b5g.P.L="Dissolved Phosphorus (ug/L",
                                         CTD.Temperature.b0C= "Temperature (deg C)")
```

```{r, warning = FALSE, meassage = FALSE, echo = FALSE}
#Create Table

summary.Erie_dropNAs %>%
gt() %>%
 


  cols_width(
    c(Variable)~ px(120),
    c(Obs)~ px(120),
    everything() ~ px(90))%>%

  
  
  tab_header(
    title = md("Summary Statistics: Lake Erie Harmful Algal Bloom Characteristics at Bottom and Surface Depths")) %>%
  
 
  fmt_passthrough (columns=c(Variable)) %>%
  fmt_number(columns = c(Obs), decimals = 0) %>%
  fmt_number(columns = c(Median), decimals = 2) %>%
  fmt_number(columns = c(Mean), decimals = 2) %>%
  fmt_number(columns = c(SD), decimals = 2) %>%
  

cols_label(
    Variable = "",
    Obs = "Oberservations",
    Median = "Median",
    Mean = "Mean",
    SD = "Standard Deviation")
  

```

***Table 2. Summary Statistics: Harmful Algal Bloom and Lake Characteristics of Lake Erie 2012 through 20187***

```{r, include = FALSE, message = FALSE, warning = FALSE}
#Add a year column
Erie_Data_2012_2018 <- Erie_Surface_2012_2018 %>% 
  mutate(Year = format(Date, "%Y"))

# Select Variables of Interest
Erie.select<-Erie_Data_2012_2018%>%
 dplyr::select(Year, Dissolved.Microcystin.b5g.L, Extracted.Chlorophyll.a.b5g.L, CTD.Dissolved.Oxygen.mg.L, Nitrate..Nitrite.mg.N.L, Total.Dissolved.Phosphorus.b5g.P.L, CTD.Temperature.b0C)
head(Erie.select)

```

```{r, include = FALSE, message = FALSE, warning = FALSE}

Erie.select.longer<-pivot_longer(Erie.select, !Year, names_to = "Variable", values_to = "value")

summary.Erie<-Erie.select.longer%>%
  dplyr::group_by(Year, Variable) %>%
  dplyr::summarize(
    Obs=length(value)-sum(is.na(value)),
    Mean=mean(value, na.rm=TRUE),
    SD=sd(value, na.rm=TRUE),
    Median=median(value, na.rm=TRUE))
   
head(summary.Erie)

#remove columns which have NAs
summary.Erie_dropNAs <- summary.Erie %>% 
  drop_na()
  
```

```{r, include = FALSE, message = FALSE, warning = FALSE}
#recode
summary.Erie_dropNAs$Variable <- dplyr:: recode(summary.Erie_dropNAs$Variable, 
                                         Dissolved.Microcystin.b5g.L = "Dissolved Microcystin (ug/L)",
                                         Extracted.Chlorophyll.a.b5g.L="Chlorophyll a (ug/L)",
                                         CTD.Dissolved.Oxygen.mg.L="Dissolved Oxygen (mg/L)",
                                         Nitrate..Nitrite.mg.N.L="Nitrate/Nitrite Concentration (mg/L)",
                                         Total.Dissolved.Phosphorus.b5g.P.L="Dissolved Phosphorus (ug/L",
                                         CTD.Temperature.b0C= "Temperature (deg C)")
```

```{r, warning = FALSE, meassage = FALSE, echo = FALSE}
#Create Table

summary.Erie_dropNAs %>%
gt() %>%
 


  cols_width(
    c(Variable)~ px(120),
    c(Obs)~ px(120),
    everything() ~ px(90))%>%

  
  
  tab_header(
    title = md("Summary Statistics: Lake Erie Harmful Algal Bloom Characteristics in 2012 through 2018")) %>%
  
 
  fmt_passthrough (columns=c(Variable)) %>%
  fmt_number(columns = c(Obs), decimals = 0) %>%
  fmt_number(columns = c(Median), decimals = 2) %>%
  fmt_number(columns = c(Mean), decimals = 2) %>%
  fmt_number(columns = c(SD), decimals = 2) %>%
  

cols_label(
    Variable = "",
    Obs = "Oberservations",
    Median = "Median",
    Mean = "Mean",
    SD = "Standard Deviation")
  

```

\newpage

# Analysis



## Question 1: Which variables (i.e., temperature, nitrate/nitrite, and/or phosphorus) contribute to microcystin production, Microcystis population growth, and dissolved oxygen?

## Question 2: How does Microcystis and microcystin concentration change over time?

## Question 3: Are there differences between toxin and algae concentrations at different lake depths?

Regarding the third question it is hypothesized that both microcystin and chlorophyll a concentrations will be greater at surface depths opposed to bottom depths in Lake Erie. These hypotheses are stated below:

***Hypotheses for microcystin concentration at surface vs. bottom depth***

Ho: Surface depth microcystin concentrations are <= bottom depth microcystin concentrations

HA: Surface depth microcystin concentrations are > bottom depth microcystin concentrations

***Hypotheses for chlorophyll a concentration at surface vs. bottom depth***

Ho: Surface depth Chlorophyll a concentrations are <= bottom depth microcystin concentrations

HA: Surface depth Chlorophyll a concentrations are > bottom depth microcystin concentrations

Before assessing these claims, a Shapiro-Wilk test for normality was conducted. The test rejected the null hypothesis which states that the data are normally distributed (W = 0.606, p-value = 2.2e-16). In result, the Wilcoxon Rank Sum test was employed to evaluate the means between surface and bottom layers. This test, rejected the null hypothesis which states that microcystin concentrations at surface depth are less than or equal to microcystin concentrations at bottom depth (W = 33075, p-value = 0.0339). The test also rejected that null hypothesis which states that chlorophyll a concentrations at surface depth are less than or equal to chlorophyll a concentrations at bottom depth (W = 152684, p-value = 3.783e-05). Both tests yielded p-values where were below 0.05. 

```{r, include = FALSE}
shapiro.test(bottom_surface_Erie$Dissolved.Microcystin.b5g.L) #the data is not shown to be normal; reject the null hypothesis of normality in this case
```


```{r, include = FALSE}
#Drop NAs

bottom_surface_Erie_dropNA_microcystin <- bottom_surface_Erie %>% 
  drop_na(Dissolved.Microcystin.b5g.L)

bottom_surface_Erie_dropNA_CHLa <- bottom_surface_Erie %>% 
  drop_na(Extracted.Chlorophyll.a.b5g.L)
```

```{r, include = FALSE}
#flip the data

bottom_surface_Erie_dropNA_microcystin$Sample.Depth.category <- factor(bottom_surface_Erie_dropNA_microcystin$Sample.Depth.category, levels = c("Surface", "Bottom"))

bottom_surface_Erie_dropNA_CHLa$Sample.Depth.category <- factor(bottom_surface_Erie_dropNA_CHLa$Sample.Depth.category, levels = c("Surface", "Bottom"))
```


```{r, include = FALSE}
#Microcystin: surface vs bottom

wilcox.test(Dissolved.Microcystin.b5g.L ~ Sample.Depth.category, data = bottom_surface_Erie_dropNA_microcystin, alternative = "greater") #Due to nonnormality of the data, a rank sum test was applied

```

```{r, include = FALSE}
#CHLa: surface vs bottom

wilcox.test(Extracted.Chlorophyll.a.b5g.L ~ Sample.Depth.category, data = bottom_surface_Erie_dropNA_CHLa, alternative = "greater") #Due to nonnormality of the data, a rank sum test was applied
```



\newpage

# Summary and Conclusions


\newpage

# References
<add references here if relevant, otherwise delete this section> 

Centers for Disease Control and Prevention (CDC; 2022). Causes and Ecosystem Impacts. https://www.cdc.gov/habs/environment.html#:~:text=Levels%20of%20nutrients%20such%20as%20phosphorus%20and%20nitrogen%20increase%20in,Sewage%20from%20people%20and%20animals

Dean, S. (2022). Harmful algal blooms in Lake Erie expected to be smaller this year, says NOAA. https://www.michigan.gov/egle/newsroom/mi-environment/2022/08/11/harmful-algal-blooms-in-lake-erie-expected-to-be-smaller-this-year-says-noaa#:~:text=The%20Lake%20Erie%20HABs%20consist,treatment%20of%20their%20source%20water.
Dierkes, C. (2014). Algae Through the Ages. https://ohioseagrant.osu.edu/news/2014/17gmc/algae-through-the-ages

Lake Erie Foundation (LEF; n.d.) Harmful Algal Blooms. https://lakeeriefoundation.org/issues/nutrientsharmful-algae/#:~:text=Harmful%20algal%20blooms%20(HABs)%20are,and%20the%20resulting%20algal%20blooms.

National Oceanic and Atmospheric Administration (NOAA; 2016). What is a harmful algal bloom? https://www.noaa.gov/what-is-harmful-algal-bloom

National Science Foundation (NSF; 2019). Lake Erie's toxic algae blooms: Why is the water turning green? https://beta.nsf.gov/news/lake-eries-toxic-algae-blooms-why-water-turning#:~:text=Since%20the%20late%201990s%2C%20Lake,warnings%20to%20avoid%20water%20contact. 

United States Environmental Protection Agency (U.S. EPA; 2022a). Harmful Algal Blooms. https://www.epa.gov/nutrientpollution/harmful-algal-blooms

United States Environmental Protection Agency (U.S. EPA; 2022b). Lake Erie. https://www.epa.gov/greatlakes/lake-erie
U.S. National Office for HABs. Recent Trends: National Changes. (n.d.) https://hab.whoi.edu/maps/regions-us-distribution/regions-us-recent-trends/
```

